### 2020-05-20 manuscript/manuscript.Rmd
updates to skeleton

---
title: 'NSync: Boy bands meet nitrogen'
author:
- Hollister. J. W. ^\*^ *^1^*, Kellogg, D. Q. *^2^*, Arnold, C.*^3^*, Gold,
  A.*^2^*, Wilson, E. H.*^3^*, Chadwick, C.*^3^*, Dickson, D.*^3^*, Forshay, K.
  J.*^4^*
- *^1^* US Environmental Protection Agency, Office Of Research and Development,
  Atlantic Coastal Environmental Sciences Division, Narragansett, RI 02882
- *^2^* University of Rhode Island, Department of Natural Resources Science,
  Kingston, RI 02881
- *^3^* University of Connecticut, Center for Land Use Education and Research,
  Haddam, CT 06438
- *^4^* US Environmental Protection Agency, Office Of Research and Development,
  Groundwater Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <hollister.jeff@epa.gov*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: Great abstract goes here
---

```{r setup, include=FALSE}
source("../R/packages.R")
opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")
options(scipen = 2, digits = 5)
```

Introduction
============

Intro goes here and citations look like [@vitousek1997human; @finlay2013human; @filippelli2008global].

Methods
=======

Methods are what belong here and an inline fig citation looks like (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`) and to reference that fig, assuming it is an image file inside the ../figures folder would look this way [@bivand2014maptools].

![Figure 1: Map of URI Watershed Watch lake and reservoir sampling sites](media/image1.jpg){width="5.7in" height="7.45in"}

`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`

If you want some inline R code, do this `r nrow(iris)`.

Results
=======

Results go here and if you want a table caption, do this (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`). And then this in an example code chunk kicing out a pander table.

```{r ww_params, echo=FALSE, results="asis"}
pander::pander(head(iris), style = "rmarkdown")
```

`r tab_num("ww_params")`

Discussion and conclusions
==========================

Really great conclusions and discussion thereof goes here!

Bibliography
============

### 2020-05-20 R/packages.R

@@ -1,0 +1,2 @@
+library(knitr)
+library(captioner)

### 2020-05-20 manuscript/manuscript.Rmd
skeleton mostly good

@@
 author:
-- Hollister. J. W. ^\*^ *^1^*, Kellogg, D. Q. *^2^*, Arnold, C.*^3^*, Gold,
+- |
+  Hollister. J. W. ^\*^ *^1^*, Kellogg, D. Q. *^2^*, Arnold, C.*^3^*, Gold,
   A.*^2^*, Wilson, E. H.*^3^*, Chadwick, C.*^3^*, Dickson, D.*^3^*, Forshay, K.
   J.*^4^*
-- *^1^* US Environmental Protection Agency, Office Of Research and Development,
+- |
+  *^1^* US Environmental Protection Agency, Office Of Research and Development,
   Atlantic Coastal Environmental Sciences Division, Narragansett, RI 02882
-- *^2^* University of Rhode Island, Department of Natural Resources Science,
+- |
+  *^2^* University of Rhode Island, Department of Natural Resources Science,
   Kingston, RI 02881
-- *^3^* University of Connecticut, Center for Land Use Education and Research,
+- |
+  *^3^* University of Connecticut, Center for Land Use Education and Research,
   Haddam, CT 06438
-- *^4^* US Environmental Protection Agency, Office Of Research and Development,
+- |
+  *^4^* US Environmental Protection Agency, Office Of Research and Development,
   Groundwater Characterization and Remediation Division, Ada, OK 74820
-- '^\*^ \*corresponding author: <hollister.jeff@epa.gov*>'
+- |
+  '^\*^ \*corresponding author: <hollister.jeff@epa.gov*>'
 bibliography: manuscript.bib
@@
 
-![Figure 1: Map of URI Watershed Watch lake and reservoir sampling sites](media/image1.jpg){width="5.7in" height="7.45in"}
-
-`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`
+![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../figures/ww_map.jpg){width="5.7in" height="7.45in"}
 

### 2020-05-28 manuscript/manuscript.Rmd
Update manuscript.Rmd
Changes to intro and methods

@@
 
+Excess nitrogen (N) delivery via surface water to coastal estuaries contributes to impaired water quality evidenced by excess algal blooms and hypoxia (citation).
+Identifying landscape N sinks (areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system) and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations.
+N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale, and makes extensive use of widely available GIS data.
+The theoretical underpinnings rely on decades of research and are outlined in [@kellogg2010geospatial].
+The landscape N sinks included in this approach include wetlands, streams and ponds, which are all areas were water flow is slowed and N transformations have time to occur.
+
+The nsink R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps.
+These static maps include 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from a grid of points across a watershed, highlighting “leaky” areas with less downstream N retention vs.
+those with higher downstream retention; and c) N Delivery Index, combining maps (a) and (b) along with source loading based on NLCD categories and expressed as an N index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream.
+
 Methods
@@
 If you want some inline R code, do this `r nrow(iris)`.
+
+Starting point - less well-developed NHD during development stage forced less versatile architecture.
+
+Adaptation - With more advanced NHD data and online SSURGO, along with increased familiarity with R gave us the opportunity to re-write, making use of better data and faster architecture using R.
+
+Changes to underlying methodology:
+
+Streams and Lakes: use NHD Plus V2 to estimate retention/travel time rather doing independent calcs. Making use of USGS expertise.
+Wetlands: we used to focus on riparian wetlands and used hydric soils as sink.
+With SSURGO we include all hydric soils, regardless of whether they abut surface water.
+Hydric status is calculated within each soil mapping unit according to the % hydric within each soil mapping unit, taking the average for that 30X30 cell.[JEFF, I need to check exactly what we do here].
+Removal is estimated as 80% X %hydric within the mapping unit.
 
@@
 ============
+@article{kellogg2010geospatial,
+  title={A geospatial approach for assessing denitrification sinks within lower-order catchments},
+  author={Kellogg, DQ and Gold, Arthur J and Cox, Suzanne and Addy, Kelly and August, Peter V},
+  journal={Ecological Engineering},
+  volume={36},
+  number={11},
+  pages={1596--1606},
+  year={2010},
+  publisher={Elsevier}
+}

### 2020-05-28 manuscript/manuscript.Rmd
updated reference in manuscript.Rmd

@@
 ============
-@article{kellogg2010geospatial,
-  title={A geospatial approach for assessing denitrification sinks within lower-order catchments},
-  author={Kellogg, DQ and Gold, Arthur J and Cox, Suzanne and Addy, Kelly and August, Peter V},
-  journal={Ecological Engineering},
-  volume={36},
-  number={11},
-  pages={1596--1606},
-  year={2010},
-  publisher={Elsevier}
-}

### 2021-06-08 manuscript/manuscript.Rmd
Added abstract, adapted vignette, fleshed out intro and segues. Added citations.

--- 2020-05-20 manuscript/manuscript.Rmd
@@
 ---
-title: 'NSync: Boy bands meet nitrogen'
+title: 'Assessing landscape N removal in coastal New England catchments using the N-Sink approach with a new R package, nsink'
+# alternative title in keeping with PeerJ: nsink: an R package for estimating and visualizing N removal from landscape N sinks
 author:
@@
 - |
-  '^\*^ \*corresponding author: <hollister.jeff@epa.gov*>'
+  '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
 bibliography: manuscript.bib
@@
   chunk_output_type: console
-csl: ecology.csl
-abstract: Great abstract goes here
+csl: peerj.csl
+abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
+Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery.
+Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system.
+Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to evaluate the effects of their decisions on downstream resources.
+N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
+A series of static maps are produced to visualize N removal efficiency, N loading, transport and delivery, the latter three in the form of indices, To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England.
+Users can use the R package to investigate a HUC-12 of interest, or can visit the UConn website to explore the already mapped areas.
+Users can also investigate specific flowpaths interactively by clicking on any location within the catchment.
+A flowpath is generated, along with a table describing N removal along each segment.
+
 ---
@@
 
-Intro goes here and citations look like [@vitousek1997human; @finlay2013human; @filippelli2008global].
+Excess nitrogen (N) delivery via surface water to coastal estuaries contributes to impaired water quality evidenced by excess algal blooms and hypoxia [@ryther1971nitrogen; @howarth2006nitrogen; @nixon1995coastal].
+Hydrologic connections and flowpaths are influential in the delivery of nutrients to estuaries [@mengistu2020wetland].
+Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations.
+The intent of N-Sink, and its associated R package `nsink`, is as a screening tool to identify areas where downgradient N sinks are able to process excess N vs.
+areas where downgradient N sinks are less prevalent or effective at removing N.
+The latter argues for more aggressive N management at the source, while the former argues for protecting and/or restoring downgradient N sinks that are helping to reduce N delivery to estuaries.
+N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale, and makes extensive use of widely available GIS data.
+The theoretical underpinnings are outlined in [@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrificaiton] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
+Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
+
+The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps.
+These static maps include 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from a grid of points across a watershed, highlighting “leaky” areas with less downstream N retention vs.
+those with higher downstream retention; and c) N Delivery Index, combining maps (a) and (b) along with source loading based on NLCD categories and expressed as an N index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream.
 
@@
 If you want some inline R code, do this `r nrow(iris)`.
+~~~~~~
+
+# Development of the R package `nsink`
+The package `nsink` implements an approach to estimate relative nitrogen (N) removal along a flow path.
+This approach is detailed in Kellogg et al.
+[-@kellogg2010geospatial].
+The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder.
+That version required time-consuming data manipulation by hand due to limitations of earlier NHD datasets as well as the limitations of working in a GIS environment requiring a user license.
+With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink could be automated and more easily applied to other HUC-12 catchments, leading to the development of the `nsink` package.
+Specifically, `nsink` relies on packages `sf` and `raster`.
+
+# Using the `nsink` package
+The `nsink` package provides several functions to set up and run an N-Sink 
+analysis for a specified 12-digit HUC code.  All required data are downloaded, 
+prepared for the analysis, HUC-wide nitrogen removal calculated, and flow paths 
+summarized.  Additionally, a convenience function that will run all of the 
+required functions for a specified HUC is included.
+
+The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Efficiency, and N Delivery Index.
+
+# Download the data
+`nsink_get_huc_id()` - search all 12-digit HUC names using a known location name.
+
+`nsink_get_data()` - download and save in a data folder hydrography, soils, and land cover for the specified 12-digit HUC. 
+These data will cover the HUC, but will not yet be reduced to those data exclusive to the specified HUC.
+The `nsink` package utilizes openly available data from several U.S. Federal 
+Government sources. All data are from publicly available sources and
+as of `r lubridate::today()` no authentication is required to access these 
+sources.  The HUC ID is required and users may specify a path for storing the 
+data as well as indicate whether or not to download the data again if they 
+already exist in the data directory.
+
+# Prepare the data
+Once the data are downloaded there are several additional data processing steps 
+that are required to subset the data to the HUC and set all data to a 
+common coordinate reference system (CRS).
+
+These include:
+
+- filter out the HUC boundary
+- mask all other data sets to the HUC boundary
+- convert all column names to lower case
+- create new columns
+- harmonize raster extents
+- set all data to common CRS
+
+The `nsink_prep_data()` function will complete all of these processing steps.  
+It requires a HUC ID, a specified CRS, and a path to a data directory.  It 
+returns a "list" (a specific data format in R) with all required data for subsequent N-Sink analyses.
+NOTE: Not sure where we are at with the CRS. Check for latest.
+A quick note on the CRS.  In the near future, the preferred way to specify the CRS values will either 
+be with Well-Known Text (WKT) or [EPSG Codes](http://spatialreference.org/ref/epsg/). 
+Proj.4 strings will eventually be deprecated. Currently the packages that `nsink` relies on are a
+t different stages in implementing the changes to PROJ. Currently Proj.4 strings are preferred, 
+but that will change soon.
+
+# Calculate N removal
+The next step in the N-Sink process is to calculate relative nitrogen removal.  
+Details on how the nitrogen removal estimates are calculated are available in 
+Kellogg et al. [-@kellogg2010geospatial]. Since the publication of that article,
+changes have been implemented to make better use of the most recently available 
+geospatial data; data that were not available at the time of publication. Those 
+changes are described in Section XX. The `nsink_calc_removal()` function 
+takes the prepared data as an input and returns a "list" with three items:
+
+- raster_method: This item contains a raster based approach to calculating 
+removal.  Used for the static maps of removal.
+- land_removal: This represents land based nitrogen removal which is hydric 
+soils with areas of impervious surface removed.
+- network_removal: This contains removal along the NHD Plus flow network.  
+Removal is calculated separately for streams and waterbodies (i.e., lakes and 
+reservoirs).
+
+# Generate and summarize flowpaths
+A useful part of the N-Sink approach is not just the calculation of relative 
+removal for individual components of the landscape, it is the ability to 
+summarize that removal along the legnth of a specified flowpath.  The `nsink` 
+package provides two functions that facilitate this process. The 
+`nsink_generate_flowpath()` function takes a point location as an `sf` object 
+and the prepped data (generated by `nsink_prep_data()`) as input and returns an 
+`sf` LINESTRING of the flowpath starting from the input point location and 
+terminating at the furthest downstream location in the input NHD Plus.  The 
+flowpath on land is generated from a flow direction grid.  Once that flowpath 
+intersects the surface water network, flow is determined by flow along the NHD Plus 
+stream network.  First, create the `sf` POINT object.
+
+```{r flowpath_start}
+# Load up the sf package
+library(sf)
+# Starting point
+pt <- c(1948121, 2295822)
+start_loc <- st_sf(st_sfc(st_point(c(pt)), crs = aea))
+```
+
+You may also determine your point location interactively by plotting your data 
+and using the `locator()` function .  First create a simple plot.
+
+```{r flowpath_plot, eval=FALSE}
+# Create a simple plot
+plot(st_geometry(niantic_data$huc))
+plot(st_geometry(niantic_data$lakes), add = T, col = "darkblue")
+plot(st_geometry(niantic_data$streams), add = T, col = "blue")
+```
+
+With the map made, you can use that to interactively select a location and use 
+the x and y to create the `sf` POINT object.
+
+```{r interactive_loc, eval=FALSE}
+# Select location on map for starting point
+pt <- unlist(locator(n = 1))
+# convert to sf POINT
+start_loc_inter <- st_sf(st_sfc(st_point(pt), crs = aea))
+```
+
+With a point identified, we can use that as the starting location for our 
+flowpath.
+
+```{r generate_flowpath, eval=FALSE}
+niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
+```
+
+The reutrned value has both the `flowpath_ends`, the portion of the flowpath on
+the land which is created using the flow direction grid, and the 
+`flowpath_network` which is the portion of the flowpath from the NHD Plus 
+network that occur after the upstream `flowpath_ends` intersect the surface water network.
+
+```{r thefp}
+niantic_fp
+```
+
+With a flowpath generated we can summarize the relative nitrogen removal along 
+that flowpath with the `nsink_summarize_flowpath()` function. It takes the 
+flowpath and removal as input.  A data frame is returned with each segment 
+identified by type, the percent removal associated with that segment, and 
+relative removal.  Total relative removal is 100 - minimum of the `n_out` 
+column. That is, `n_out` keeps track of the proportion of N leaving each 
+subsequent downstream segment.
+
+```{r summarize_it}
+niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal)
+niantic_fp_removal
+100-min(niantic_fp_removal$n_out)
+```
+
+# Static maps
+
+Individual flow paths are useful for specific areas of interest, but it is also
+useful to look at removal patterns across the landscape.  The 
+`nsink_generate_static_maps()` function provides these HUC-wide rasters.  
+Required inputs are the prepped data, removal raster, and sampling density, for which a default is provided.   
+The function returns four separate rasters.
+
+- `removal_effic`: Landscape wide estimate of relative nitrogen removal as a
+percentage.
+- `loading_idx`: An index of relative nitrogen loads by land cover class derived 
+from published sources.
+- `transport_idx`: This is relative nitrogen transport for a sample of all 
+possible flowpaths in a given HUC.  This is an expensive computational task, so 
+`nsink` generates a removal hotspot map based on a sample of flowpaths and the 
+final hotspot map is interpolated from these samples and referred to as the 
+nitrogen transport index.  The `samp_density` argument controls the number of 
+sample flowpaths generated.  
+- `delivery_idx`: The delivery index is the combination of the loading index and 
+the transport index  It represents which areas of the landscape are 
+delivering the most nitrogen to the outflow of the watershed.
+
+```{r static_maps, eval=FALSE}
+niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 
+                                                  900)
+```
+
+And with these static maps generated, you can plot them quickly with 
+`nsink_plot_removal()`, `nsink_plot_delivery()`, or `nsink_plot_transport()`.  
+An example of `nsink_plot_transport()` is below.
+
+```{r plots}
+nsink_plot_transport(niantic_static_maps$transport_idx)
+```
+
+# Convenience function: Build it all!
+
+The workflow described above includes all the basic functionality.  Some users 
+may wish to use `nsink` to calculate the base layers for an N-Sink analysis and 
+then build an application outside of R.  A convenience function that downloads 
+all data, prepares, calculates removal, and generates static maps has been 
+included to facilitate this type of analysis.  The `nsink_build()` function 
+requires a HUC ID, coordinate reference system, and sampling density.  An output
+folder is also needed but has a default location.  There are ptional arguments for 
+forcing a new download and playing a sound to signal when the build has 
+finished. All prepped data files and `.tif` files are 
+written to the output folder for use in other applications.
+
+```{r build, eval=FALSE}
+niantic_huc <- nsink_get_huc_id("Niantic River")$huc_12
+aea <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 
++ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
+nsink_build(niantic_huc, aea, samp_dens = 900)
+```
+
+Currently we have created the static maps for all HUC-12s in coastal southern New England (map). These may be viewed and explored interactively at https://XXXX.
+
+# Changes to underlying methodology:
+
+Several improvements have been made to the underlying methodology that have allowed `nsink` to make use of more recently available spatial data that had to be estimated in the original version.
+In estimating N removal in streams, we originally had to estimated stream depth and velocity based on other available geospatial data to arrive at a travel time along a given stream reach.
+With the updated and expanded NHD Plus V2 we can use the estimates provided in that dataset, making use of USGS expertise that went into these estimates.
+
+In estimating N removal occurring in the terrestrial portion of the flowpath, we originally focused on riparian wetlands, using SSURGO mapped hydric soils to identify this landscape sink.
+With the latest version of SSURGO we include all hydric soils in the catchment, except those identified as "subaqueous".
+Each soil mapping unit (SMU) that has a positive hydric status also has an estimate of %hydric for that SMU.
+We then estimate N removal within that SMU as 80% of of the %hydric.
+Each raster has one SMU identified with it and raster cell size is 30m X 30m, which is the maximum wetland width originally estimated as having 80% N removal.
+In the current version N removal is cumulative as a flowpath intercepts hydric soils in any cell before encountering surface water.
+
+# Off-network hydrology
+In the NHD Plus dataset there are situations where a water body or stream segment is "off-network", i.e., it is not linked to the larger surface flow network.
+These may be features such as groundwater fed kettle ponds, or canals or ditches where connecting hydrology may not be captured in the NHD Plus dataset.
+These off-network features tend to have less NHD spatial data associated with them, making direct N removal estimates impossible.
+In these cases, we must estimate removal based on other in-network surface water features.
+While off-network features are rare and spatially small compared to the overall system, they do occur in some catchments.
+We estimated removal from off-network stream segments as the median removal of all first order streams in the same catchment.
+In this case we are assuming these first order segments behave similarly to other first order stream reaches.
+Off-network ditches and canals, identified as such in the NHD dataset, are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore assign removal as the lower quartile of removal from the highest order streams in the same catchment.
+Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment [@hampton2019residence; ].
+
+# Caveats
+N-Sink was designed for catchments with a surface water network, so is not intended for use in urbanized areas where surface water flow is highly manipulated and the unit of interest is a sewershed vs.
+a watershed.
+N removal from landscape sinks in urban areas would be impossible to predict.
+Similarly, if agricultural areas have a high density of tile drains, N removal estimates would not reflect the lower residence time and associated removal.
+
+Here we present two examples using the static maps generated by the `nsink` package to illustrate how we envision this tool being useful in decision-making at the local level.
+
+Use Examples
+===========================================================================================
+N-Sink is now widely available for HUC-12 catchments in coastal CT, RI and southern MA and decision-makers are only now beginning to make use of maps and data available online (ADD URL).
+However, we are aware of various ways this tool is starting to be integrated into risk analyses.
+Here we demonstrate two examples using the Niantic River (CT) the Farm River (CT) HUCs.
 

### 2021-06-17 manuscript/manuscript.Rmd
Edits before sending to Chet. This is the version I sent.

@@
 author:
-- |
-  Hollister. J. W. ^\*^ *^1^*, Kellogg, D. Q. *^2^*, Arnold, C.*^3^*, Gold,
-  A.*^2^*, Wilson, E. H.*^3^*, Chadwick, C.*^3^*, Dickson, D.*^3^*, Forshay, K.
-  J.*^4^*
-- |
-  *^1^* US Environmental Protection Agency, Office Of Research and Development,
+- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Arnold, C.^*3*^, Gold,
+  A.^*2*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K.
+  J.^*4*^
+- ^*1*^ US Environmental Protection Agency, Office Of Research and Development,
   Atlantic Coastal Environmental Sciences Division, Narragansett, RI 02882
-- |
-  *^2^* University of Rhode Island, Department of Natural Resources Science,
+- ^*2*^ University of Rhode Island, Department of Natural Resources Science,
   Kingston, RI 02881
-- |
-  *^3^* University of Connecticut, Center for Land Use Education and Research,
+- ^*3*^ University of Connecticut, Center for Land Use Education and Research,
   Haddam, CT 06438
-- |
-  *^4^* US Environmental Protection Agency, Office Of Research and Development,
+- ^*4*^ US Environmental Protection Agency, Office Of Research and Development,
   Groundwater Characterization and Remediation Division, Ada, OK 74820
-- |
-  '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
+- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
 bibliography: manuscript.bib
@@
   chunk_output_type: console
-csl: peerj.csl
-abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
+csl: ecology.csl
+abstract: 'Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
 Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery.
@@
 Users can also investigate specific flowpaths interactively by clicking on any location within the catchment.
-A flowpath is generated, along with a table describing N removal along each segment.
-
+A flowpath is generated, along with a table describing N removal along each segment.'
 ---
 
-```{r setup, include=FALSE}
-source("../R/packages.R")
-opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
-tab_num <- captioner(prefix = "Table")
-fig_num <- captioner(prefix = "Figure")
-options(scipen = 2, digits = 5)
+```{r, include = FALSE}
+knitr::opts_chunk$set(
+  eval = TRUE,
+  cache = FALSE,
+  tidy = FALSE
+)
+load(system.file("testdata.rda", package = "nsink"))
+```
+
+```{r setup, echo=FALSE}
+library(nsink)
 ```
@@
 N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale, and makes extensive use of widely available GIS data.
-The theoretical underpinnings are outlined in [@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrificaiton] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
+The theoretical underpinnings are outlined in [@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
 Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
@@
 
-Methods are what belong here and an inline fig citation looks like (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`) and to reference that fig, assuming it is an image file inside the ../figures folder would look this way [@bivand2014maptools].
-
-![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../figures/ww_map.jpg){width="5.7in" height="7.45in"}
-
-If you want some inline R code, do this `r nrow(iris)`.
-~~~~~~
-
-# Development of the R package `nsink`
+## Development of the R package `nsink`
 The package `nsink` implements an approach to estimate relative nitrogen (N) removal along a flow path.
@@
 
-# Using the `nsink` package
+## Using the `nsink` package
 The `nsink` package provides several functions to set up and run an N-Sink 
@@
 
-# Download the data
+## Download the data
 `nsink_get_huc_id()` - search all 12-digit HUC names using a known location name.
@@
 
-# Prepare the data
+## Prepare the data
 Once the data are downloaded there are several additional data processing steps 
@@
 
-# Calculate N removal
+## Calculate N removal
 The next step in the N-Sink process is to calculate relative nitrogen removal.  
@@
 
-# Generate and summarize flowpaths
+## Generate and summarize flowpaths
 A useful part of the N-Sink approach is not just the calculation of relative 
@@
 
-# Static maps
+## Static maps
 
@@
 useful to look at removal patterns across the landscape.  The 
-`nsink_generate_static_maps()` function provides these HUC-wide rasters.  
-Required inputs are the prepped data, removal raster, and sampling density, for which a default is provided.   
+`nsink_generate_static_maps()` function provides these HUC-wide rasters. Required inputs are the prepped data, removal raster, and sampling density, for which a default is provided.   
 The function returns four separate rasters.
@@
 
-# Convenience function: Build it all!
+## Convenience function: Build it all!
 
@@
 
-Currently we have created the static maps for all HUC-12s in coastal southern New England (map). These may be viewed and explored interactively at https://XXXX.
+Currently we have created the static maps for all HUC-12s in coastal southern New England (map). These may be viewed and explored interactively at https://clear.uconn.edu/projects/nsink/.
 
-# Changes to underlying methodology:
+## Changes to underlying methodology:
 
@@
 
-# Off-network hydrology
+## Off-network hydrology
 In the NHD Plus dataset there are situations where a water body or stream segment is "off-network", i.e., it is not linked to the larger surface flow network.
@@
 
-# Caveats
-N-Sink was designed for catchments with a surface water network, so is not intended for use in urbanized areas where surface water flow is highly manipulated and the unit of interest is a sewershed vs.
-a watershed.
-N removal from landscape sinks in urban areas would be impossible to predict.
-Similarly, if agricultural areas have a high density of tile drains, N removal estimates would not reflect the lower residence time and associated removal.
+## Caveats
+N-Sink was designed for catchments with a surface water network.
+It is not intended in arid areas or where surface water flow is highly manipulated, such as in urbanized or tile-drained agricultural areas.
+In these situations surface water flow is highly variable and often difficult to estimate or predict.
 
@@
 Use Examples
-===========================================================================================
-N-Sink is now widely available for HUC-12 catchments in coastal CT, RI and southern MA and decision-makers are only now beginning to make use of maps and data available online (ADD URL).
-However, we are aware of various ways this tool is starting to be integrated into risk analyses.
-Here we demonstrate two examples using the Niantic River (CT) the Farm River (CT) HUCs.
+============
+N-Sink is now available for HUC-12 catchments in coastal CT, RI and southern MA and decision-makers are beginning to make use of maps and data available as an [interactive tool](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31).
+However, we are aware of various ways this tool is starting to be integrated into risk analyses. Here we demonstrate two examples using the Niantic River (CT) the Farm River (CT) HUCs.
 
@@
 =======
-
-Results go here and if you want a table caption, do this (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
-And then this in an example code chunk kicing out a pander table.
-
-```{r ww_params, echo=FALSE, results="asis"}
-pander::pander(head(iris), style = "rmarkdown")
-```
-
-`r tab_num("ww_params")`
 

### 2022-02-22 manuscript/manuscript.Rmd
Lots of changes -- progress!

@@
 ---
-title: 'Assessing landscape N removal in coastal New England catchments using the N-Sink approach with a new R package, nsink'
-# alternative title in keeping with PeerJ: nsink: an R package for estimating and visualizing N removal from landscape N sinks
+title: Assessing landscape N removal in coastal New England catchments using the N-Sink
+  approach with a new R package, nsink
 author:
-- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Arnold, C.^*3*^, Gold,
-  A.^*2*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K.
-  J.^*4*^
-- ^*1*^ US Environmental Protection Agency, Office Of Research and Development,
-  Atlantic Coastal Environmental Sciences Division, Narragansett, RI 02882
-- ^*2*^ University of Rhode Island, Department of Natural Resources Science,
-  Kingston, RI 02881
-- ^*3*^ University of Connecticut, Center for Land Use Education and Research,
-  Haddam, CT 06438
-- ^*4*^ US Environmental Protection Agency, Office Of Research and Development,
-  Groundwater Characterization and Remediation Division, Ada, OK 74820
+- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Arnold, C.^*3*^, Gold, A.^*2*^,
+  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
+- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
+  Coastal Environmental Sciences Division, Narragansett, RI 02882
+- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
+  RI 02881
+- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
+  CT 06438
+- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
+  Characterization and Remediation Division, Ada, OK 74820
 - '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
@@
 output:
+  html_document:
+    df_print: paged
   word_document:
@@
 csl: ecology.csl
-abstract: 'Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
-Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery.
-Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system.
-Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to evaluate the effects of their decisions on downstream resources.
-N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
-A series of static maps are produced to visualize N removal efficiency, N loading, transport and delivery, the latter three in the form of indices, To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England.
-Users can use the R package to investigate a HUC-12 of interest, or can visit the UConn website to explore the already mapped areas.
-Users can also investigate specific flowpaths interactively by clicking on any location within the catchment.
-A flowpath is generated, along with a table describing N removal along each segment.'
+abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water
+  quality. Land management decisions made within estuarine watersheds have a direct
+  impact on downstream N delivery. Natural features within watersheds can act as landscape
+  sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous
+  N, effectively removing it from the aquatic system. Identifying and evaluating these
+  landscape sinks and their spatial relationship to N sources can assist managers
+  to evaluate the effects of their decisions on downstream resources. N-Sink is an
+  approach that uses widely available GIS data to identify landscape sinks within
+  HUC-12 catchments, estimate their N removal potential and map the effect of those
+  sinks on N movement through the catchment. A series of static maps are produced
+  to visualize N removal efficiency, N loading, transport and delivery, the latter
+  three in the form of indices, To facilitate data acquisition, processing and visualization,
+  an R package `nsink` was developed and has been used to produce static maps for
+  all HUC-12 catchments in southern New England. Users can use the R package to investigate
+  a HUC-12 of interest, or can visit the UConn website to explore the already mapped
+  areas. Users can also investigate specific flowpaths interactively by clicking on
+  any location within the catchment. A flowpath is generated, along with a table describing
+  N removal along each segment.
 ---
@@
   cache = FALSE,
-  tidy = FALSE
+  tidy = FALSE,
+  message = FALSE
 )
-load(system.file("testdata.rda", package = "nsink"))
+#load(system.file("testdata.rda", package = "nsink"))
 ```
 
-```{r setup, echo=FALSE}
+```{r setup, echo=FALSE, warning = FALSE}
+
 library(nsink)
+```
+
+```{r get_huc, include = FALSE, eval = TRUE}
+# The Niantic
+niantic_huc_id <- nsink_get_huc_id("Niantic River")$huc_12
+# The Farm
+farm_huc_id <- nsink_get_huc_id("Farm River")$huc_12
+```
+
+```{r get_data_norun, eval = FALSE, include = FALSE}
+# Change eval = TRUE if you want to download the data
+# Get data for selected HUC
+niantic_download <- nsink_get_data(niantic_huc_id, 
+                                   data_dir = "nsink_niantic_data")
+farm_download <- nsink_get_data(farm_huc_id, data_dir = "nsink_farm_data")
 ```
@@
 ## Development of the R package `nsink`
-The package `nsink` implements an approach to estimate relative nitrogen (N) removal along a flow path.
-This approach is detailed in Kellogg et al.
-[-@kellogg2010geospatial].
+The package `nsink` implements the approach detailed in Kellogg et al.
+[-@kellogg2010geospatial] to estimate relative nitrogen (N) removal along a flowpath.
 The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder.
@@
 
+Functions|Description
+:--------|:----------
+nsink_get_huc_id|Takes Hydrologic Unit Code (HUC) Name and returns HUC12 IDs
+nsink_get_data|Downloads local copies of datasets for specified HUC
+nsink_prep_data|Standardizes projections, clips all datasets to HUC boundary
+nsink_calc_removal|Calculates percentage of nitrogen (N) removal within the various landscape N sinks
+nsink_generate_flowpath|Generates a flowpath from XY starting point for use in N removal analysis
+nsink_summarize_flowpath|Estimates N removal along flowpath and lists removal by type of N sink encountered
+nsink_generate_static_maps|Generates static maps for a given HUC
+nsink_plot|Creates simple plot with pre-selected palettes from a list of static maps created by nsink_generate_static_maps()
+nsink_build|Wrapper function that runs all required functions to build the full dataset and static maps for a specified HUC12
+
 ## Using the `nsink` package
-The `nsink` package provides several functions to set up and run an N-Sink 
-analysis for a specified 12-digit HUC code.  All required data are downloaded, 
-prepared for the analysis, HUC-wide nitrogen removal calculated, and flow paths 
-summarized.  Additionally, a convenience function that will run all of the 
-required functions for a specified HUC is included.
+The `nsink` package provides several functions to set up and run an N-Sink analysis for a specified 12-digit HUC.
+All required data are downloaded, prepared for the analysis, HUC-wide nitrogen removal calculated, and flowpaths summarized.
+Additionally, a convenience function that will run all of the required functions for a specified HUC is included.
 
@@
 ## Download the data
-`nsink_get_huc_id()` - search all 12-digit HUC names using a known location name.
+`nsink_get_huc_id()` - search 12-digit HUC IDs using a known location name.
 
 `nsink_get_data()` - download and save in a data folder hydrography, soils, and land cover for the specified 12-digit HUC.
-These data will cover the HUC, but will not yet be reduced to those data exclusive to the specified HUC.
-The `nsink` package utilizes openly available data from several U.S. Federal 
-Government sources. All data are from publicly available sources and
-as of `r lubridate::today()` no authentication is required to access these 
-sources.  The HUC ID is required and users may specify a path for storing the 
-data as well as indicate whether or not to download the data again if they 
-already exist in the data directory.
+These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC.
+The `nsink` package utilizes openly available data from several U.S.
+Federal Government sources and as of `r lubridate::today()` no authentication is required to access these sources.
+The HUC ID is required and users may specify a path for storing the data as well as indicate whether or not to download the data again if they already exist in the data directory.
 
 ## Prepare the data
-Once the data are downloaded there are several additional data processing steps 
-that are required to subset the data to the HUC and set all data to a 
-common coordinate reference system (CRS).
+Once the data are downloaded there are several additional data processing steps that are required to subset the data to the HUC and set all data to a common coordinate reference system (CRS).
 
@@
 The `nsink_prep_data()` function will complete all of these processing steps.
-It requires a HUC ID, a specified CRS, and a path to a data directory.  It 
-returns a "list" (a specific data format in R) with all required data for subsequent N-Sink analyses.
-NOTE: Not sure where we are at with the CRS. Check for latest.
-A quick note on the CRS.  In the near future, the preferred way to specify the CRS values will either 
-be with Well-Known Text (WKT) or [EPSG Codes](http://spatialreference.org/ref/epsg/). 
-Proj.4 strings will eventually be deprecated. Currently the packages that `nsink` relies on are a
-t different stages in implementing the changes to PROJ. Currently Proj.4 strings are preferred, 
-but that will change soon.
+It requires a HUC ID, a specified CRS, and a path to a data directory.
+It returns a "list" (a specific data format in R) with all required data for subsequent N-Sink analyses.
+
+```{r prep_data, eval = TRUE, include = FALSE}
+# EPSG for CONUS Albers Equal Area
+aea <- 5072
+
+# Prep data for selected HUC
+niantic_data <- nsink_prep_data(niantic_huc_id, projection = aea, data_dir = "nsink_niantic_data")
+farm_data <- nsink_prep_data(farm_huc_id, projection = aea,
+                             data_dir = "nsink_farm_data")
+```
 
 ## Calculate N removal
-The next step in the N-Sink process is to calculate relative nitrogen removal.  
-Details on how the nitrogen removal estimates are calculated are available in 
-Kellogg et al. [-@kellogg2010geospatial]. Since the publication of that article,
-changes have been implemented to make better use of the most recently available 
-geospatial data; data that were not available at the time of publication. Those 
-changes are described in Section XX. The `nsink_calc_removal()` function 
-takes the prepared data as an input and returns a "list" with three items:
+The next step in the N-Sink workflow is to calculate potential nitrogen removal within each landscape sink, expressed as percent of incoming N.
+Details on how the nitrogen removal estimates are calculated are available in Kellogg et al.
+[-@kellogg2010geospatial], and make use peer-reviewed literature.
+Since the publication of that article, changes have been implemented to make better use of the most recently available geospatial data.
+Those changes are described below.
+The `nsink_calc_removal()` function takes the prepared data as an input and returns a "list" with three items:
 
-- raster_method: This item contains a raster based approach to calculating 
-removal.  Used for the static maps of removal.
-- land_removal: This represents land based nitrogen removal which is hydric 
-soils with areas of impervious surface removed.
-- network_removal: This contains removal along the NHD Plus flow network.  
-Removal is calculated separately for streams and waterbodies (i.e., lakes and 
-reservoirs).
+- raster_method: Contains a raster-based approach to calculating removal.  Used for the static maps showing potential N removal within landscape sinks.
+- land_removal: Represents land-based nitrogen removal within vegetated hydric soils (with impervious surface, i.e., developed areas, removed).
+- network_removal: Contains removal along the NHD Plus flow network. Removal is calculated separately for streams and waterbodies (i.e., lakes and reservoirs).
+
+```{r calc_removal, eval = TRUE, include = FALSE}
+# Calculate removal from prepped data
+niantic_removal <- nsink_calc_removal(niantic_data)
+```
 
 ## Generate and summarize flowpaths
-A useful part of the N-Sink approach is not just the calculation of relative 
-removal for individual components of the landscape, it is the ability to 
-summarize that removal along the legnth of a specified flowpath.  The `nsink` 
-package provides two functions that facilitate this process. The 
-`nsink_generate_flowpath()` function takes a point location as an `sf` object 
-and the prepped data (generated by `nsink_prep_data()`) as input and returns an 
-`sf` LINESTRING of the flowpath starting from the input point location and 
-terminating at the furthest downstream location in the input NHD Plus.  The 
-flowpath on land is generated from a flow direction grid.  Once that flowpath 
-intersects the surface water network, flow is determined by flow along the NHD Plus 
-stream network.  First, create the `sf` POINT object.
+A useful part of the N-Sink approach is not just the calculation of potential N removal for individual components of the landscape, it is the ability to summarize cumulative removal along the length of a specified flowpath.
+Starting from any specified location the flowpath on land is generated from a flow direction grid.
+Once that flowpath intersects the surface water network, flow is determined by flow along the NHD Plus surface water network.
+With a flowpath generated the cumulative N removal along that flowpath can be calculated with the `nsink_summarize_flowpath()` function.
+It takes the flowpath and removal as input.
+A data frame is returned with each segment identified by type, the percent removal associated with that segment, and cumulative removal.
+Total cumulative removal is 100 - minimum of the `n_out` column.
+That is, `n_out` keeps track of the percent of N leaving each subsequent downstream segment (Table 2).
 
-```{r flowpath_start}
-# Load up the sf package
+```{r summarize_it, message = FALSE, include = FALSE}
 library(sf)
+## Linking to GEOS 3.8.1, GDAL 3.1.4, PROJ 6.3.1
 # Starting point
@@
 start_loc <- st_sf(st_sfc(st_point(c(pt)), crs = aea))
-```
-
-You may also determine your point location interactively by plotting your data 
-and using the `locator()` function .  First create a simple plot.
-
-```{r flowpath_plot, eval=FALSE}
-# Create a simple plot
-plot(st_geometry(niantic_data$huc))
-plot(st_geometry(niantic_data$lakes), add = T, col = "darkblue")
-plot(st_geometry(niantic_data$streams), add = T, col = "blue")
-```
-
-With the map made, you can use that to interactively select a location and use 
-the x and y to create the `sf` POINT object.
-
-```{r interactive_loc, eval=FALSE}
-# Select location on map for starting point
-pt <- unlist(locator(n = 1))
-# convert to sf POINT
-start_loc_inter <- st_sf(st_sfc(st_point(pt), crs = aea))
-```
-
-With a point identified, we can use that as the starting location for our 
-flowpath.
-
-```{r generate_flowpath, eval=FALSE}
 niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
-```
-
-The reutrned value has both the `flowpath_ends`, the portion of the flowpath on
-the land which is created using the flow direction grid, and the 
-`flowpath_network` which is the portion of the flowpath from the NHD Plus 
-network that occur after the upstream `flowpath_ends` intersect the surface water network.
-
-```{r thefp}
-niantic_fp
-```
-
-With a flowpath generated we can summarize the relative nitrogen removal along 
-that flowpath with the `nsink_summarize_flowpath()` function. It takes the 
-flowpath and removal as input.  A data frame is returned with each segment 
-identified by type, the percent removal associated with that segment, and 
-relative removal.  Total relative removal is 100 - minimum of the `n_out` 
-column. That is, `n_out` keeps track of the proportion of N leaving each 
-subsequent downstream segment.
-
-```{r summarize_it}
 niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal)
@@
 ```
-
+```{r show_summary, include = TRUE}
+kable(niantic_fp_removal, caption = "Summary table of N removal along specified flowpath")
+```
 ## Static maps
 
-Individual flow paths are useful for specific areas of interest, but it is also
-useful to look at removal patterns across the landscape.  The 
-`nsink_generate_static_maps()` function provides these HUC-wide rasters. Required inputs are the prepped data, removal raster, and sampling density, for which a default is provided.   
+Individual flowpaths are useful for specific areas of interest, but it is also useful to look at removal patterns across the landscape.
+The `nsink_generate_static_maps()` function produces a set of HUC-wide raster maps.
+Required inputs are the prepped data, removal raster, and sampling density, for which a default is provided.
 The function returns four separate rasters.
 
-- `removal_effic`: Landscape wide estimate of relative nitrogen removal as a
-percentage.
-- `loading_idx`: An index of relative nitrogen loads by land cover class derived 
-from published sources.
-- `transport_idx`: This is relative nitrogen transport for a sample of all 
-possible flowpaths in a given HUC.  This is an expensive computational task, so 
-`nsink` generates a removal hotspot map based on a sample of flowpaths and the 
-final hotspot map is interpolated from these samples and referred to as the 
-nitrogen transport index.  The `samp_density` argument controls the number of 
-sample flowpaths generated.  
-- `delivery_idx`: The delivery index is the combination of the loading index and 
-the transport index  It represents which areas of the landscape are 
-delivering the most nitrogen to the outflow of the watershed.
+- `removal_effic`: HUC-wide estimate of potential nitrogen removal from different landscape sinks as a percentage.
+- `loading_idx`: An normalized index of nitrogen loads by land cover class derived from published sources, ranging from 0 to 1.
+- `transport_idx`: N transport for a sample of all possible flowpaths in a given HUC.
+This is an expensive computational task, so `nsink` generates a removal hotspot map based on a sample of flowpaths and the final hotspot map is interpolated from these samples and referred to as the nitrogen transport index.
+The `samp_density` argument controls the number of sample flowpaths generated.
+Each starting point is assigned the value of N delivery (%) based on cumulative N removal along the flowpath from that starting point.
+- `delivery_idx`: The delivery index is the combination of the loading index and the transport index  It highlights areas of the landscape that are delivering the most nitrogen to the outflow of the watershed based on NLCD land cover data.
 
-```{r static_maps, eval=FALSE}
-niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 
-                                                  900)
+```{r static_maps, eval=TRUE, include = FALSE}
+niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 900)
 ```
 
-And with these static maps generated, you can plot them quickly with 
-`nsink_plot_removal()`, `nsink_plot_delivery()`, or `nsink_plot_transport()`.  
-An example of `nsink_plot_transport()` is below.
+These static maps can be quickly plotted with `nsink_plot()` using the `map` argument to identify map type.
 
-```{r plots}
-nsink_plot_transport(niantic_static_maps$transport_idx)
+```{r plots, fig.cap = "Nitrogen Removal Efficiency (NRE, range 0 to 100%) is the estimeated percentage of N removed within three types of landscape N sinks - wetlands, streams, ponds."}
+
+nsink_plot(niantic_static_maps, map = "removal")
+#nsink_plot(niantic_static_maps, map = "transport")
+#nsink_plot(niantic_static_maps, map = "delivery")
+
 ```
@@
 
-The workflow described above includes all the basic functionality.  Some users 
-may wish to use `nsink` to calculate the base layers for an N-Sink analysis and 
-then build an application outside of R.  A convenience function that downloads 
-all data, prepares, calculates removal, and generates static maps has been 
-included to facilitate this type of analysis.  The `nsink_build()` function 
-requires a HUC ID, coordinate reference system, and sampling density.  An output
-folder is also needed but has a default location.  There are ptional arguments for 
-forcing a new download and playing a sound to signal when the build has 
-finished. All prepped data files and `.tif` files are 
-written to the output folder for use in other applications.
+The workflow described above includes all the basic functionality.
+Some users may wish to use `nsink` to calculate the base layers for an N-Sink analysis and then build an application outside of R.
+A convenience function that downloads all data, prepares, calculates removal, and generates static maps has been included to facilitate this type of analysis.
+The `nsink_build()` function requires a HUC ID, coordinate reference system, and sampling density.
+An output folder is also needed but has a default location.
+There are optional arguments for forcing a new download and playing a sound to signal when the build has finished.
+All prepped data files and `.tif` files are written to the output folder for use in other applications.
 
-```{r build, eval=FALSE}
-niantic_huc <- nsink_get_huc_id("Niantic River")$huc_12
-aea <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 
-+ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
-nsink_build(niantic_huc, aea, samp_dens = 900)
-```
-
-Currently we have created the static maps for all HUC-12s in coastal southern New England (map). These may be viewed and explored interactively at https://clear.uconn.edu/projects/nsink/.
-
-## Changes to underlying methodology:
+## Changes to underlying methodology
 
@@
 While off-network features are rare and spatially small compared to the overall system, they do occur in some catchments.
-We estimated removal from off-network stream segments as the median removal of all first order streams in the same catchment.
+We estimated removal from off-network stream segments to be equal to the median removal of all first order streams in the same catchment.
 In this case we are assuming these first order segments behave similarly to other first order stream reaches.

### 2022-03-08 manuscript/manuscript.Rmd
Cleaned up text, added figures, tables, and use examples.

@@
 # Starting point
-pt <- c(1948121, 2295822)
-start_loc <- st_sf(st_sfc(st_point(c(pt)), crs = aea))
-niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
-niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal)
-niantic_fp_removal
-100-min(niantic_fp_removal$n_out)
+
+pta <- c(1953529, 2290811)
+start_loc_inter_a <- st_sf(st_sfc(st_point(pta), crs = aea))
+niantic_fp_a <- nsink_generate_flowpath(start_loc_inter_a, niantic_data)
+niantic_fp_removal_a <- nsink_summarize_flowpath(niantic_fp_a, niantic_removal)
+
 ```
-```{r show_summary, include = TRUE}
-kable(niantic_fp_removal, caption = "Summary table of N removal along specified flowpath")
+```{r show_summary, include = TRUE, echo = FALSE}
+knitr::kable(niantic_fp_removal_a, caption = "Table 1: Summary of N removal along specified flowpath A")
 ```
@@
 
-```{r plots, fig.cap = "Nitrogen Removal Efficiency (NRE, range 0 to 100%) is the estimeated percentage of N removed within three types of landscape N sinks - wetlands, streams, ponds."}
+```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (NRE, range 0 to 100%) is the estimeated percentage of N removed within three types of landscape N sinks - wetlands, streams, ponds."}
 
 nsink_plot(niantic_static_maps, map = "removal")
-#nsink_plot(niantic_static_maps, map = "transport")
-#nsink_plot(niantic_static_maps, map = "delivery")
+```
+```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (NTI, range 0 to 100%) is the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment."}
 
+nsink_plot(niantic_static_maps, map = "transport")
 ```
+```{r plots3, fig.cap = "Figure 3: Nitrogen Delivery Index (NDI, range 0 to 100%)  is a relative measure of the estimated N being transported from a location within a mapped watershed to receiving waters.
+N loading estimates are based on land cover classes, making use of peer-reviewed published data, and normalized, with a range of 0 to 1.
+These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the NDI."}
 
-## Convenience function: Build it all!
+nsink_plot(niantic_static_maps, map = "delivery")
+```
+## Convenience function to build it all
 
@@
 All prepped data files and `.tif` files are written to the output folder for use in other applications.
+
+N-Sink maps are now available for HUC-12 catchments in coastal CT, RI and southern MA as a web application [interactive tool](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31).
+This app can be used to examine and better understand N movement through these HUC-12 catchments.
 
@@
 Each soil mapping unit (SMU) that has a positive hydric status also has an estimate of %hydric for that SMU.
-We then estimate N removal within that SMU as 80% of of the %hydric.
+We then estimate N removal within that SMU as 80% of the %hydric.
 Each raster has one SMU identified with it and raster cell size is 30m X 30m, which is the maximum wetland width originally estimated as having 80% N removal.
@@
 Off-network ditches and canals, identified as such in the NHD dataset, are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore assign removal as the lower quartile of removal from the highest order streams in the same catchment.
-Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment [@hampton2019residence; ].
+Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment [@hampton2019residence].
 
 ## Caveats
-N-Sink was designed for catchments with a surface water network.
-It is not intended in arid areas or where surface water flow is highly manipulated, such as in urbanized or tile-drained agricultural areas.
-In these situations surface water flow is highly variable and often difficult to estimate or predict.
-
-Here we present two examples using the static maps generated by the `nsink` package to illustrate how we envision this tool being useful in decision-making at the local level.
+N-Sink was designed for catchments where surface and groundwater flow paths are not highly manipulated.
+In settings with substantial storm water conveyance networks or with extensive subsurface agricultural drainage, the removal of N in riparian zones is likely to be overstated by the model.
+The numeric outputs of the tool focus on metrics suitable for planning purposes, such as percent removal of N from source to receiving water and relative intensity of loading from different locations.
+N-Sink is a decision support tool and was created to be a useful, easy way for local land use managers to explore the relationship between land use and nitrogen pollution in their waters.
+N-Sink uses the best available science on land use/nitrogen interactions, plus widely available datasets for waterway networks, soils and land use, to highlight major sources and sinks of nitrogen within a watershed context.
 
@@
 ============
-N-Sink is now available for HUC-12 catchments in coastal CT, RI and southern MA and decision-makers are beginning to make use of maps and data available as an [interactive tool](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31).
-However, we are aware of various ways this tool is starting to be integrated into risk analyses. Here we demonstrate two examples using the Niantic River (CT) the Farm River (CT) HUCs.
+### Example 1: Comparison of N delivery from two adjacent locations
+Comparing the N delivery from two locations that are separated by approximately 900 meters illustrates the importance of location with respect to landscape N sinks.
+Approximately 56% of the N that originates at point A is transported to the estuary vs.
+approximately 25% of the N that originates at point B (Fig 4).
+The difference is due to the different N sinks encountered along the way to the outlet or estuary.
+Most of the removal that occurs along flowpath A is due to flow through hydric (wetland) soils, with little removal occurring during flow along the downgradient stream network (Table 2).
+Flowpath B encounters two ponds along the way, significantly increasing residence time and N transformation (Table 3).
 
-Results
-=======
+```{r plots4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations with differing N removal estimates.
+Hydrofeatures are pictured on the NTI (Nitrongen Transport Index) map."}
+
+ptb <- c(1954039, 2291592)
+start_loc_inter_b <- st_sf(st_sfc(st_point(ptb), crs = aea))
+
+niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
+niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)
+raster::plot(niantic_static_maps$transport_idx, breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), 
+             col = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", "#E2F088", "#F5E871",
+                     "#F9C159", "#F99844", "#F56A2E","#EF2820"),
+                 main = "Flowpaths originating at points A and B, separated by ~ 900 m")
+plot(niantic_data$streams, add = TRUE, col = "dodgerblue")
+plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
+plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray50", lwd = 3.0)
+plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray50", lwd = 3.0)
+plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
+plot(st_geometry(niantic_fp_b$flowpath_ends), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
+x <- c(pta[1], ptb[1])
+y <- c(pta[2], ptb[2])
+ab <- st_multipoint(matrix(c(x,y), nrow = 2, ncol = 2))
+ab_plot <- st_sf(st_sfc(ab), crs = aea)
+plot(ab_plot, add = TRUE, pch = 16)
+text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)
+
+```
+```{r compare_flowpaths, include = TRUE, echo = FALSE}
+knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B")
+```
+
+### Example 2: Source controls and sink protection or restoration
+The maps produced by nsink can be helpful in prioritizing source controls, sink protection and/or restoration.
+While wetlands are considered "protected" by law they are frequently altered through variances granted by local commissions or committees charged with zoning decisions.
+By better understanding the role sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving water.
+Variances granted on a case=by-case basis can, however unintentionally, quickly undermine local water quality goals.
+
+Similarly, sources of N can be highlighted using the nsink static maps [Fig XX]. The N Delivery Index map combines N sources with N transport to highlight
+
+Potential ways to use N-Sink are also available as [bonus maps](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31) on the N-Sink website.
 

### 2022-03-14 manuscript/manuscript.Rmd
Version sent to Jeff, Art and Chet

@@
 output:
+  word_document:
+    reference_docx: manuscript_template.docx
   html_document:
     df_print: paged
-  word_document:
-    reference_docx: manuscript_template.docx
 editor_options:
@@
 ```{r setup, echo=FALSE, warning = FALSE}
-
+library(FedData)
 library(nsink)
+library(raster)
 ```
@@
 Specifically, `nsink` relies on packages `sf` and `raster`.
+The `nsink` package is available from https://github.com/usepa/nsink and is fully described in [@hollister2022nsink].
 
@@
 ```{r static_maps, eval=TRUE, include = FALSE}
-niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 900)
+niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 450)
 ```
@@
 ### Example 1: Comparison of N delivery from two adjacent locations
-Comparing the N delivery from two locations that are separated by approximately 900 meters illustrates the importance of location with respect to landscape N sinks.
+Users can use the following to interactively identify a starting point for a flowpath of interest. 
+```{r case1, echo = TRUE, eval = FALSE}
+# EPSG for CONUS Albers Equal Area
+aea <- 5072
+# Select location on map for starting point
+pt <- unlist(locator(n = 1))
+# Convert to sf POINT
+start_loc <- st_sf(st_sfc(st_point(pt), crs = aea))
+# Generate the flowpath
+niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
+# Summarize N removal along the flowpath
+niantic_removal <- nsink_calc_removal(niantic_data)
+niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal) # table of removal within flowpath segments
+```
+
+Using this approach, we can compare the N delivery from two locations that are separated by approximately 900 meters.
+This example illustrates the importance of location with respect to landscape N sinks.
 Approximately 56% of the N that originates at point A is transported to the estuary vs.
-approximately 25% of the N that originates at point B (Fig 4).
-The difference is due to the different N sinks encountered along the way to the outlet or estuary.
+approximately 25% of the N that originates at point B (Figure 4).
+The difference is due to the different N sinks encountered along the way to the estuary.
 Most of the removal that occurs along flowpath A is due to flow through hydric (wetland) soils, with little removal occurring during flow along the downgradient stream network (Table 2).
-Flowpath B encounters two ponds along the way, significantly increasing residence time and N transformation (Table 3).
+Flowpath B encounters a large pond along the way, significantly increasing residence time and N transformation (Table 3).
+These two undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet.
+If development were proposed in the western portion of this area, source controls could be advised, while the less "leaky "eastern region underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing N transformation.
 
-```{r plots4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations with differing N removal estimates.
+```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations with differing N removal estimates.
 Hydrofeatures are pictured on the NTI (Nitrongen Transport Index) map."}
@@
 text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)
+```
 
-```
 ```{r compare_flowpaths, include = TRUE, echo = FALSE}
@@
 ### Example 2: Source controls and sink protection or restoration
-The maps produced by nsink can be helpful in prioritizing source controls, sink protection and/or restoration.
-While wetlands are considered "protected" by law they are frequently altered through variances granted by local commissions or committees charged with zoning decisions.
-By better understanding the role sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving water.
-Variances granted on a case=by-case basis can, however unintentionally, quickly undermine local water quality goals.
+The maps produced by N-sink can be helpful in prioritizing source controls, sink protection and/or restoration.
+For example, wetlands are considered "protected" by law but are frequently altered through variances granted by local commissions or committees charged with zoning decisions.
+By better understanding the role N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters.
+Variances granted on a case=by-case basis can quickly undermine local water quality goals.
+Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize acquisition or wetland restoration projects.
 
-Similarly, sources of N can be highlighted using the nsink static maps [Fig XX]. The N Delivery Index map combines N sources with N transport to highlight
+Similarly, sources of N that are contributing relatively more N due to their location within the catchment can be highlighted using the `nsink` static maps.
+The N Delivery Index (NDI) map combines N sources with N transport to highlight developed areas that are located on "leakier parts" of a catchment.
+Figure 5A shows land cover in a smaller area of the watershed, with two developed areas circled.
+Figure 5B shows the N Delivery Index (NDI) from these two developed areas.
+Note that the northern area is less densely developed than the southern area but the NDI is higher in the northern area as compared to the southern area, suggesting that N source controls within the northern development would be worth investigating.
 
-Potential ways to use N-Sink are also available as [bonus maps](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31) on the N-Sink website.
+```{r figure5a, include = TRUE, echo = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest, highlighting two developed areas."}
+box <- c(1949902, 1953930, 2288202, 2292262)
+e <- extent(box)
+legend_nlcd <- pal_nlcd()
+case2 <- crop(niantic_data$nlcd, e)
+vals <- unique(case2@data@values)
+leg <- legend_nlcd[legend_nlcd$code %in% vals,]
+case2_north <- st_sf(st_sfc(st_point(c(1950846, 2290912)), crs = aea)) 
+case2_south <- st_sf(st_sfc(st_point(c(1952693, 2289249)), crs = aea))
 
-Discussion and conclusions
+my_key <- list(rectangles=list(col = leg$color),
+              text=list(lab=leg$description),
+              space='left',
+              columns=1,
+              size=2,
+              cex=.6)
+plot(case2, col = legend_nlcd$color)
+plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
+plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
+```
+```{r, figure5b, include = TRUE, echo = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index (NDI) over land cover, with hotter colors denoting higher N delivery as estimated based on normalized loading from different land covers multiplied by the N Transport Efficiency (NTE)."}
+case2_delivery <- crop(niantic_static_maps$delivery_idx, e)
+plot(case2, col = legend_nlcd$color, alpha = 0.6, legend = FALSE)
+plot(case2_delivery, breaks = c(20, 40, 60, 80, 100), col = c("#FFBEBE", "#F57A7A","#A80405","#652600"), add = TRUE)
+plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
+plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
+
+```
+
+Other examples of N-Sink are also available as [bonus maps](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31) on the N-Sink website.
+
+Conclusions
 ==========================
 
-Really great conclusions and discussion thereof goes here!
+N-Sink as implemented in the `nsink` R package allows users to easily download all necessary spatial data (hydrography, land cover and soils), create static maps (N Removal Efficiency, N Transport Efficiency and N Delivery Index), and interactively investigate N removal along flowpaths within any specified catchment.
+Currently, the default catchment definition uses HUC-12 designations, though it is also possible to use larger catchments.
+Larger catchments would require longer download and processing times.
+For investigators who do not work in R, we have created an interactive web application that covers HUC-12 catchments in coastal CT, RI and southern MA [interactive tool](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31).
+
+Local decision-makers that have prioritized nitrogen mitigation in their long-term planning efforts can use this tool to help them better understand the potential impact of proposed development projects and zoning variances.
+Similarly, land trusts and other NGOs interested in N mitigation can use this tool to identify high priority areas for acquisition or restoration.
 

### 2022-04-26 manuscript/manuscript.Rmd
Merged comments/edits from Chet and Art; added citations

@@
 author:
-- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Arnold, C.^*3*^, Gold, A.^*2*^,
+- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
   Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
-- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
+- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
+  RI 02881
+- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
   Coastal Environmental Sciences Division, Narragansett, RI 02882
-- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
-  RI 02881
 - ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
@@
 output:
+  html_document:
+    df_print: paged
   word_document:
     reference_docx: manuscript_template.docx
-  html_document:
-    df_print: paged
 editor_options:
@@
 csl: ecology.csl
-abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water
-  quality. Land management decisions made within estuarine watersheds have a direct
-  impact on downstream N delivery. Natural features within watersheds can act as landscape
-  sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous
-  N, effectively removing it from the aquatic system. Identifying and evaluating these
-  landscape sinks and their spatial relationship to N sources can assist managers
-  to evaluate the effects of their decisions on downstream resources. N-Sink is an
-  approach that uses widely available GIS data to identify landscape sinks within
-  HUC-12 catchments, estimate their N removal potential and map the effect of those
-  sinks on N movement through the catchment. A series of static maps are produced
-  to visualize N removal efficiency, N loading, transport and delivery, the latter
-  three in the form of indices, To facilitate data acquisition, processing and visualization,
-  an R package `nsink` was developed and has been used to produce static maps for
-  all HUC-12 catchments in southern New England. Users can use the R package to investigate
-  a HUC-12 of interest, or can visit the UConn website to explore the already mapped
-  areas. Users can also investigate specific flowpaths interactively by clicking on
-  any location within the catchment. A flowpath is generated, along with a table describing
-  N removal along each segment.
+abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
+Land management decisions made within estuarine watersheds have a directimpact on downstream N delivery.
+Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system.
+Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to evaluate the effects of their decisions on downstream resources.
+N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
+A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index.
+To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England.
+Users can use the R package to develop N delivery maps for a HUC-12 of interest, or can visit the University of Connecticut website to explore the already mapped areas.
+Users can also investigate specific flowpaths interactively by clicking on any location within the catchment.
+A flowpath is generated, along with a table describing N removal along each segment.
 ---
@@
 Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations.
-The intent of N-Sink, and its associated R package `nsink`, is as a screening tool to identify areas where downgradient N sinks are able to process excess N vs.
-areas where downgradient N sinks are less prevalent or effective at removing N.
+The intent of N-Sink, and its associated R package `nsink`, is as a screening tool to identify source areas where downgradient N sinks are able to process and remove excess N, versus areas where downgradient N sinks are less prevalent or effective at removing N.
 The latter argues for more aggressive N management at the source, while the former argues for protecting and/or restoring downgradient N sinks that are helping to reduce N delivery to estuaries.
-N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale, and makes extensive use of widely available GIS data.
-The theoretical underpinnings are outlined in [@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
+
+N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data.
+The theoretical underpinnings are outlined in Kellogg et al.
+[-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
 Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
 
-The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps.
-These static maps include 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from a grid of points across a watershed, highlighting “leaky” areas with less downstream N retention vs.
-those with higher downstream retention; and c) N Delivery Index, combining maps (a) and (b) along with source loading based on NLCD categories and expressed as an N index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream.
+The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps.Datasets used by `nsink` are the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web].
+The static maps include 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed.
+This heat map highlights “leaky” source areas with less downstream N retention versus those with higher downstream retention; and c) N Delivery Index, combining maps (1) and (2) along with source loading based on NLCD categories and expressed as an index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream.
 
@@
 The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder.
-That version required time-consuming data manipulation by hand due to limitations of earlier NHD datasets as well as the limitations of working in a GIS environment requiring a user license.
-With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink could be automated and more easily applied to other HUC-12 catchments, leading to the development of the `nsink` package.
+That version required time-consuming data manipulation by hand due to limitations of earlier NHDPlus datasets as well as the limitations of working in a GIS environment requiring a user license.
+With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink was automated and applied to other HUC-12 catchments, leading to the development of the `nsink` package.
 Specifically, `nsink` relies on packages `sf` and `raster`.
@@
 The `nsink` package provides several functions to set up and run an N-Sink analysis for a specified 12-digit HUC.
-All required data are downloaded, prepared for the analysis, HUC-wide nitrogen removal calculated, and flowpaths summarized.
-Additionally, a convenience function that will run all of the required functions for a specified HUC is included.
+By using the package all required data are downloaded, prepared for the analysis, HUC-wide N removal calculated, and flowpaths summarized.
+Additionally, a convenience function is included that will run all of the required functions for a specified HUC.
 
@@
 Details on how the nitrogen removal estimates are calculated are available in Kellogg et al.
-[-@kellogg2010geospatial], and make use peer-reviewed literature.
+[-@kellogg2010geospatial], and make use of peer-reviewed literature.
 Since the publication of that article, changes have been implemented to make better use of the most recently available geospatial data.
@@
 It takes the flowpath and removal as input.
-A data frame is returned with each segment identified by type, the percent removal associated with that segment, and cumulative removal.
+A data frame is returned with each segment identified by type of sink, if present (i.e., "Hydric", "Stream", "Lake/Pond", or "No Removal"), the percent removal associated with that segment, and cumulative removal.
 Total cumulative removal is 100 - minimum of the `n_out` column.
@@
 
-pta <- c(1953529, 2290811)
+pta <- c(1954039, 2291592)
 start_loc_inter_a <- st_sf(st_sfc(st_point(pta), crs = aea))
@@
 ```{r show_summary, include = TRUE, echo = FALSE}
-knitr::kable(niantic_fp_removal_a, caption = "Table 1: Summary of N removal along specified flowpath A")
+knitr::kable(niantic_fp_removal_a, caption = "Table 1: Summary of N removal along specified flowpath A (See Figure 4)")
 ```
@@
 Each starting point is assigned the value of N delivery (%) based on cumulative N removal along the flowpath from that starting point.
-- `delivery_idx`: The delivery index is the combination of the loading index and the transport index  It highlights areas of the landscape that are delivering the most nitrogen to the outflow of the watershed based on NLCD land cover data.
+A sampling density of 300 was used to create the static maps in the web application, resulting in a roughly 30 m resolution.
+- `delivery_idx`: The delivery index is the combination of the loading index and the transport index  It highlights areas of the landscape that are delivering the most nitrogen to the outflow of the watershed based on NLCD land cover data and transport efficiency.
 
@@
 
-```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (NRE, range 0 to 100%) is the estimeated percentage of N removed within three types of landscape N sinks - wetlands, streams, ponds."}
+```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 100%) is the estimeated percentage of N removed within three types of landscape N sinks - wetlands, streams, ponds."}
 
@@
 ```
-```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (NTI, range 0 to 100%) is the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment."}
+```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (range 0 to 100%) is the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment."}
 
@@
 ```
-```{r plots3, fig.cap = "Figure 3: Nitrogen Delivery Index (NDI, range 0 to 100%)  is a relative measure of the estimated N being transported from a location within a mapped watershed to receiving waters.
+```{r plot3, fig.cap = "Figure 3: Nitrogen Delivery Index (range 0 to 100%)  is a relative measure of the estimated N being transported from a location within a mapped watershed to receiving waters.
 N loading estimates are based on land cover classes, making use of peer-reviewed published data, and normalized, with a range of 0 to 1.
@@
 ```
+
 ## Convenience function to build it all
@@
 
-Several improvements have been made to the underlying methodology that have allowed `nsink` to make use of more recently available spatial data that had to be estimated in the original version.
+Several improvements have been made to the underlying N-Sink methodology used in Kellogg et al. [-@kellogg2010geospatial] that
+have allowed `nsink` to make use of more recently available spatial data.
 In estimating N removal in streams, we originally had to estimated stream depth and velocity based on other available geospatial data to arrive at a travel time along a given stream reach.
-With the updated and expanded NHD Plus V2 we can use the estimates provided in that dataset, making use of USGS expertise that went into these estimates.
+With the updated and expanded NHD Plus V2 we use the estimates provided in that dataset, making use of USGS expertise that went into these estimates.
 
@@
 We then estimate N removal within that SMU as 80% of the %hydric.
-Each raster has one SMU identified with it and raster cell size is 30m X 30m, which is the maximum wetland width originally estimated as having 80% N removal.
+Each raster has one SMU identified with it and raster cell size is 30m X 30m.
+A 30-m wide wetland is estimated by Mayer et al.
+[-@mayer2007meta] as having 80% N removal.
 In the current version N removal is cumulative as a flowpath intercepts hydric soils in any cell before encountering surface water.
@@
 In these cases, we must estimate removal based on other in-network surface water features.
-While off-network features are rare and spatially small compared to the overall system, they do occur in some catchments.
+For the vast majority of HUC-12 catchments off-network features are rare and spatially small compared to the overall system.
+However, they can represent more than 5% of the drainage area in xx% of the catchment analyzed for southern New England.
 We estimated removal from off-network stream segments to be equal to the median removal of all first order streams in the same catchment.
 In this case we are assuming these first order segments behave similarly to other first order stream reaches.
-Off-network ditches and canals, identified as such in the NHD dataset, are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore assign removal as the lower quartile of removal from the highest order streams in the same catchment.
-Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment [@hampton2019residence].
+A subset of these off-network stream segments are identified by the NHD dataset as ditches and canals.
+These are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore N-Sink assigns removal as the lower quartile of removal from the highest order streams in the same catchment.
+Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment due to the level of uncertainty regarding residence time in these groundwater-fed lakes and ponds [@hampton2019residence].
 
@@
 N-Sink was designed for catchments where surface and groundwater flow paths are not highly manipulated.
-In settings with substantial storm water conveyance networks or with extensive subsurface agricultural drainage, the removal of N in riparian zones is likely to be overstated by the model.
-The numeric outputs of the tool focus on metrics suitable for planning purposes, such as percent removal of N from source to receiving water and relative intensity of loading from different locations.
-N-Sink is a decision support tool and was created to be a useful, easy way for local land use managers to explore the relationship between land use and nitrogen pollution in their waters.
-N-Sink uses the best available science on land use/nitrogen interactions, plus widely available datasets for waterway networks, soils and land use, to highlight major sources and sinks of nitrogen within a watershed context.
+In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage bypass the processes within wetland ecosystems, the removal of N in riparian zones is likely to be overstated.
+The variation in loading associated with different agricultural practices, density of unsewered development and inputs from wastewater treatment plants are not encompassed in the default loading index.
+Thus the numeric outputs of the tool focus on metrics suitable for planning purposes, such as percent removal of N from source to receiving water and relative intensity of loading from different locations.
+It is not intended to replace high-resolution models that rely on long term weather records and incorporate detailed spatial data on N loading, within-field transport or variable source hydrology (e.g., Easton [-@easton2008re]).
+Instead, N-Sink combines widely available datasets for waterway networks, soils and land cover with nitrogen dynamics developed for these datasets to highlight major sources and sinks of nitrogen within a watershed context.
 
@@
 ============
-### Example 1: Comparison of N delivery from two adjacent locations
+## Example 1: Comparison of N delivery from two adjacent locations
 Users can use the following to interactively identify a starting point for a flowpath of interest. 
@@
 These two undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet.
-If development were proposed in the western portion of this area, source controls could be advised, while the less "leaky "eastern region underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing N transformation.
+If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky "eastern region (flowpath B) underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing N transformation.
 
 ```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations with differing N removal estimates.
-Hydrofeatures are pictured on the NTI (Nitrongen Transport Index) map."}
+Hydrofeatures are pictured on the  Nitrongen Transport Index map."}
 
-ptb <- c(1954039, 2291592)
+ptb <- c(1953529, 2290811)
 start_loc_inter_b <- st_sf(st_sfc(st_point(ptb), crs = aea))
@@
              col = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", "#E2F088", "#F5E871",
-                     "#F9C159", "#F99844", "#F56A2E","#EF2820"),
-                 main = "Flowpaths originating at points A and B, separated by ~ 900 m")
+                     "#F9C159", "#F99844", "#F56A2E","#EF2820"))
 plot(niantic_data$streams, add = TRUE, col = "dodgerblue")
 plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
-plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray50", lwd = 3.0)
-plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray50", lwd = 3.0)
+plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray40", lwd = 3.0)
+plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray40", lwd = 3.0)
 plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
@@
 
-### Example 2: Source controls and sink protection or restoration
-The maps produced by N-sink can be helpful in prioritizing source controls, sink protection and/or restoration.
-For example, wetlands are considered "protected" by law but are frequently altered through variances granted by local commissions or committees charged with zoning decisions.
-By better understanding the role N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters.
+## Example 2: Source controls and sink protection or restoration
+The maps produced by N-Sink can be helpful in prioritizing source controls, sink protection and/or restoration.
+For example, many wetlands are considered "protected" by federal, state or local laws and regulations; but as occurs with many types of protected areas [@neal2018coastal] they are frequently altered through variances granted by local commissions or committees charged with zoning decisions.
+By better understanding the important role that N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters.
 Variances granted on a case=by-case basis can quickly undermine local water quality goals.
@@
 Similarly, sources of N that are contributing relatively more N due to their location within the catchment can be highlighted using the `nsink` static maps.
-The N Delivery Index (NDI) map combines N sources with N transport to highlight developed areas that are located on "leakier parts" of a catchment.
-Figure 5A shows land cover in a smaller area of the watershed, with two developed areas circled.
-Figure 5B shows the N Delivery Index (NDI) from these two developed areas.
-Note that the northern area is less densely developed than the southern area but the NDI is higher in the northern area as compared to the southern area, suggesting that N source controls within the northern development would be worth investigating.
+The N Delivery Index map combines N sources with N transport to highlight developed areas that are located on "leakier parts" of a catchment.
+Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled.
+Figure 5B shows the N Delivery Index from these two developed areas.
+Note that the northern area is less densely developed than the southern area but the N Delivery Index is higher in the northern area as compared to the southern area, suggesting that N source controls within the northern development might be considered.
 
-```{r figure5a, include = TRUE, echo = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest, highlighting two developed areas."}
+```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest, highlighting two developed areas."}
 box <- c(1949902, 1953930, 2288202, 2292262)
@@
 vals <- unique(case2@data@values)
-leg <- legend_nlcd[legend_nlcd$code %in% vals,]
+leg <- legend_nlcd[legend_nlcd$ID %in% vals,]
 case2_north <- st_sf(st_sfc(st_point(c(1950846, 2290912)), crs = aea)) 
@@
 
-my_key <- list(rectangles=list(col = leg$color),
-              text=list(lab=leg$description),
+my_key <- list(rectangles=list(col = leg$Color),
+              text=list(lab=leg$Class),
               space='left',
@@
               cex=.6)
-plot(case2, col = legend_nlcd$color)
+colors <- legend_nlcd$Color
+plot(case2, col = legend_nlcd$Color)
 plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
@@
 ```
-```{r, figure5b, include = TRUE, echo = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index (NDI) over land cover, with hotter colors denoting higher N delivery as estimated based on normalized loading from different land covers multiplied by the N Transport Efficiency (NTE)."}
+```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index over land cover, with hotter colors denoting higher N delivery as estimated based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
 case2_delivery <- crop(niantic_static_maps$delivery_idx, e)
-plot(case2, col = legend_nlcd$color, alpha = 0.6, legend = FALSE)
+plot(case2, col = legend_nlcd$Color, alpha = 0.6, legend = FALSE)
 plot(case2_delivery, breaks = c(20, 40, 60, 80, 100), col = c("#FFBEBE", "#F57A7A","#A80405","#652600"), add = TRUE)
@@
 
-Other examples of N-Sink are also available as [bonus maps](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31) on the N-Sink website.
+Other examples using N-Sink to examine the N loading implications of past land use decisions are also available as [bonus maps](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31) on the N-Sink website.
 
@@
 Larger catchments would require longer download and processing times.
-For investigators who do not work in R, we have created an interactive web application that covers HUC-12 catchments in coastal CT, RI and southern MA [interactive tool](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31).
+For investigators who do not work in R, we have created a web application that covers HUC-12 catchments in coastal CT, RI and southern MA using an [interactive tool](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31).
 
+N-Sink is a decision support tool and was created to be a useful, easy way for local land use managers to explore the relationship between land use and nitrogen pollution in their waters.
 Local decision-makers that have prioritized nitrogen mitigation in their long-term planning efforts can use this tool to help them better understand the potential impact of proposed development projects and zoning variances.

### 2022-04-28 manuscript/manuscript.Rmd

@@
 abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
-Land management decisions made within estuarine watersheds have a directimpact on downstream N delivery.
+Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery.
 Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system.
 Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to evaluate the effects of their decisions on downstream resources.
-N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
+N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
 A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index.
@@
 
-```{r get_data_norun, eval = FALSE, include = FALSE}
+```{r get_data_norun, eval = TRUE, include = FALSE, cache=TRUE}
 # Change eval = TRUE if you want to download the data
@@
 
-The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps.Datasets used by `nsink` are the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web].
+The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] .
+Datasets used by `nsink` are the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web].
 The static maps include 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed.
@@
 By using the package all required data are downloaded, prepared for the analysis, HUC-wide N removal calculated, and flowpaths summarized.
-Additionally, a convenience function is included that will run all of the required functions for a specified HUC.
+Additionally, a convenience function, `nsink_build()`, is included that will run all of the required functions for a specified HUC.
 
@@
 - raster_method: Contains a raster-based approach to calculating removal.  Used for the static maps showing potential N removal within landscape sinks.
-- land_removal: Represents land-based nitrogen removal within vegetated hydric soils (with impervious surface, i.e., developed areas, removed).
+- land_removal: Represents land-based nitrogen removal within vegetated hydric soils (i.e., with impervious surface/developed areas removed).
 - network_removal: Contains removal along the NHD Plus flow network. Removal is calculated separately for streams and waterbodies (i.e., lakes and reservoirs).
@@
 This is an expensive computational task, so `nsink` generates a removal hotspot map based on a sample of flowpaths and the final hotspot map is interpolated from these samples and referred to as the nitrogen transport index.
-The `samp_density` argument controls the number of sample flowpaths generated.
-Each starting point is assigned the value of N delivery (%) based on cumulative N removal along the flowpath from that starting point.
-A sampling density of 300 was used to create the static maps in the web application, resulting in a roughly 30 m resolution.
-- `delivery_idx`: The delivery index is the combination of the loading index and the transport index  It highlights areas of the landscape that are delivering the most nitrogen to the outflow of the watershed based on NLCD land cover data and transport efficiency.
+The `samp_density` argument controls the number of sample flowpaths generated and is roughly the distance between points in a uniform grid.
+Each of these starting points is assigned the value of N delivery (%) based on cumulative N removal along the flowpath from that starting point.
+The points are then interpolated with an inverse distance weighted interpolation to a roughly 30 meter resolution.
+A sampling density of 300 was used to create the static maps in the web application.
+- `delivery_idx`: The delivery index is the combination of the loading index and the transport index.
+It highlights areas of the landscape that are delivering the most nitrogen to the outflow of the watershed based on NLCD land cover data and transport efficiency.
 
@@
 An output folder is also needed but has a default location.
-There are optional arguments for forcing a new download and playing a sound to signal when the build has finished.
-All prepped data files and `.tif` files are written to the output folder for use in other applications.
+There is an optional argument for forcing a new download.
+All prepped data shape files and `.tif` files are written to the output folder for use in other applications.
 
@@
 These may be features such as groundwater fed kettle ponds, or canals or ditches where connecting hydrology may not be captured in the NHD Plus dataset.
-These off-network features tend to have less NHD spatial data associated with them, making direct N removal estimates impossible.
+These off-network features tend to have less NHD data (e.g.
+time of travel) associated with them, making direct N removal estimates impossible.
 In these cases, we must estimate removal based on other in-network surface water features.
@@
 ============
+
 ## Example 1: Comparison of N delivery from two adjacent locations
+
 Users can use the following to interactively identify a starting point for a flowpath of interest. 
@@
 ## Example 2: Source controls and sink protection or restoration
+
 The maps produced by N-Sink can be helpful in prioritizing source controls, sink protection and/or restoration.
@@
 By better understanding the important role that N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters.
-Variances granted on a case=by-case basis can quickly undermine local water quality goals.
+Variances granted on a case-by-case basis can quickly undermine local water quality goals.
 Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize acquisition or wetland restoration projects.

### 2023-04-18 manuscript/manuscript.Rmd
Incorporate edits from AG and CA

@@
 Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system.
-Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to evaluate the effects of their decisions on downstream resources.
+Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to understand the effects of their decisions on downstream resources.
 N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
@@
 To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England.
-Users can use the R package to develop N delivery maps for a HUC-12 of interest, or can visit the University of Connecticut website to explore the already mapped areas.
+The R package can be used to develop N delivery maps for a HUC-12 of interest, or users can visit the University of Connecticut website to explore the already mapped areas.
 Users can also investigate specific flowpaths interactively by clicking on any location within the catchment.
@@
 #load(system.file("testdata.rda", package = "nsink"))
+#load("testdata.rda")
 ```
@@
 
-```{r get_huc, include = FALSE, eval = TRUE}
-# The Niantic
+```{r get-huc, include = FALSE, eval = TRUE}
+# The Niantic River
 niantic_huc_id <- nsink_get_huc_id("Niantic River")$huc_12
-# The Farm
-farm_huc_id <- nsink_get_huc_id("Farm River")$huc_12
 ```
 
-```{r get_data_norun, eval = TRUE, include = FALSE, cache=TRUE}
-# Change eval = TRUE if you want to download the data
+```{r get-data, eval=TRUE, cache=FALSE, include=FALSE}
+# eval = TRUE if you want to download the data
 # Get data for selected HUC
@@
                                   data_dir = "nsink_niantic_data")
-farm_download <- nsink_get_data(farm_huc_id, data_dir = "nsink_farm_data")
 ```
@@
 The theoretical underpinnings are outlined in Kellogg et al.
-[-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
+[-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
 Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
@@
 
-Functions|Description
-:--------|:----------
-nsink_get_huc_id|Takes Hydrologic Unit Code (HUC) Name and returns HUC12 IDs
-nsink_get_data|Downloads local copies of datasets for specified HUC
-nsink_prep_data|Standardizes projections, clips all datasets to HUC boundary
-nsink_calc_removal|Calculates percentage of nitrogen (N) removal within the various landscape N sinks
-nsink_generate_flowpath|Generates a flowpath from XY starting point for use in N removal analysis
-nsink_summarize_flowpath|Estimates N removal along flowpath and lists removal by type of N sink encountered
-nsink_generate_static_maps|Generates static maps for a given HUC
-nsink_plot|Creates simple plot with pre-selected palettes from a list of static maps created by nsink_generate_static_maps()
-nsink_build|Wrapper function that runs all required functions to build the full dataset and static maps for a specified HUC12
-
 ## Using the `nsink` package
-The `nsink` package provides several functions to set up and run an N-Sink analysis for a specified 12-digit HUC.
+The `nsink` package provides several functions (Table 1) to set up and run an N-Sink analysis for a specified 12-digit HUC.
 By using the package all required data are downloaded, prepared for the analysis, HUC-wide N removal calculated, and flowpaths summarized.
 Additionally, a convenience function, `nsink_build()`, is included that will run all of the required functions for a specified HUC.
+The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Efficiency, and N Delivery Index.
 
-The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Efficiency, and N Delivery Index.
+```{r table1-build, eval = TRUE, include = FALSE}
+table1 <- data.frame(`Function` = c("nsink_get_huc_id", "nsink_get_data", 
+                                  "nsink_prep_data", "nsink_calc_removal",
+                                  "nsink_generate_flowpath", "nsink_summarize_flowpath",
+                                  "nsink_generate_static_maps", "nsink_plot",
+                                  "nsink_build"),
+                     `Description` = c("Takes Hydrologic Unit Code (HUC) Name and returns HUC12 IDs",
+                                       "Downloads local copies of datasets for specified HUC",
+                                       "Standardizes projections, clips all datasets to HUC boundary",
+                                       "Calculates percentage of nitrogen (N) removal within the various landscape N sinks",
+                                       "Generates a flowpath from XY starting point for use in N removal analysis",
+                                       "Estimates N removal along flowpath and lists removal by type of N sink encountered",
+                                       "Generates static maps for a given HUC",
+                                       "Creates simple plot with pre-selected palettes from a list of static maps created by nsink_generate_static_maps()",
+                                       "Wrapper function that runs all required functions to build the full dataset and static maps for a specified HUC12"))
+
+```
+```{r table1, include = TRUE, echo = FALSE}
+knitr::kable(table1, caption = "Table 1: Functions in the R package `nsink`")
+```
 
@@
 These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC.
-The `nsink` package utilizes openly available data from several U.S.
+The `nsink` package utilizes widely available data from several U.S.
 Federal Government sources and as of `r lubridate::today()` no authentication is required to access these sources.
@@
 
-```{r prep_data, eval = TRUE, include = FALSE}
+```{r prep-data, eval = TRUE, include = FALSE}
 # EPSG for CONUS Albers Equal Area
@@
 # Prep data for selected HUC
-niantic_data <- nsink_prep_data(niantic_huc_id, projection = aea, data_dir = "nsink_niantic_data")
-farm_data <- nsink_prep_data(farm_huc_id, projection = aea,
-                             data_dir = "nsink_farm_data")
+niantic_data <- nsink_prep_data(niantic_huc_id, projection = aea,
+                                data_dir = "nsink_niantic_data")
 ```
@@
 
-```{r calc_removal, eval = TRUE, include = FALSE}
+```{r calc-removal, eval = TRUE, include = FALSE}
 # Calculate removal from prepped data
@@
 Once that flowpath intersects the surface water network, flow is determined by flow along the NHD Plus surface water network.
-With a flowpath generated the cumulative N removal along that flowpath can be calculated with the `nsink_summarize_flowpath()` function.
-It takes the flowpath and removal as input.
+With a flowpath generated the cumulative N removal along that flowpath can be calculated with the `nsink_summarize_flowpath()` function, taking the flowpath and removal as input.
 A data frame is returned with each segment identified by type of sink, if present (i.e., "Hydric", "Stream", "Lake/Pond", or "No Removal"), the percent removal associated with that segment, and cumulative removal.
@@
 
-```{r summarize_it, message = FALSE, include = FALSE}
+```{r summarize-it, message = FALSE, include = FALSE}
 library(sf)
@@
 ```
-```{r show_summary, include = TRUE, echo = FALSE}
-knitr::kable(niantic_fp_removal_a, caption = "Table 1: Summary of N removal along specified flowpath A (See Figure 4)")
+```{r show-summary, include = TRUE, echo = FALSE}
+knitr::kable(niantic_fp_removal_a, caption = "Table 2: Summary of N removal along specified flowpath A (See Figure 4)")
 ```
@@
 
-```{r static_maps, eval=TRUE, include = FALSE}
+```{r static-maps, eval=TRUE, include = FALSE}
 niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 450)
@@
 
-```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 100%) is the estimeated percentage of N removed within three types of landscape N sinks - wetlands, streams, ponds."}
+```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 1) for the Niantic River HUC-12, showing the estimated percentage of N removed within three types of landscape N sinks - wetlands, streams, lakes/ponds/reservoirs."}
 
@@
 ```
-```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (range 0 to 100%) is the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment."}
+```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (range 0 to 100%) for the Niantic River HUC-12, showing the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment."}
 
@@
 ```
-```{r plot3, fig.cap = "Figure 3: Nitrogen Delivery Index (range 0 to 100%)  is a relative measure of the estimated N being transported from a location within a mapped watershed to receiving waters.
+```{r plot3, fig.cap = "Figure 3: Nitrogen Delivery Index (range 0 to 100%) for the Niantic River HUC-12, showing a relative measure of the estimated N being transported from a location within a mapped watershed to receiving waters.
 N loading estimates are based on land cover classes, making use of peer-reviewed published data, and normalized, with a range of 0 to 1.
@@
 We then estimate N removal within that SMU as 80% of the %hydric.
-Each raster has one SMU identified with it and raster cell size is 30m X 30m.
+Each raster has one SMU associated with it and raster cell size is 30m X 30m.
 A 30-m wide wetland is estimated by Mayer et al.
@@
 For the vast majority of HUC-12 catchments off-network features are rare and spatially small compared to the overall system.
-However, they can represent more than 5% of the drainage area in xx% of the catchment analyzed for southern New England.
+We examined seven HUC-12s within southern New England, with a range of land cover and soils, for the extent of off-network hydrology.
+None had off network stream or canal segments, while six of the seven had off-network waterbodies.
+These off-network waterbodies represented from 2 to 13% of total lake area.
 We estimated removal from off-network stream segments to be equal to the median removal of all first order streams in the same catchment.
@@
 N-Sink was designed for catchments where surface and groundwater flow paths are not highly manipulated.
-In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage bypass the processes within wetland ecosystems, the removal of N in riparian zones is likely to be overstated.
+In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage bypasses the processes within wetland ecosystems, the removal of N in riparian zones is likely to be overstated.
 The variation in loading associated with different agricultural practices, density of unsewered development and inputs from wastewater treatment plants are not encompassed in the default loading index.
@@
 aea <- 5072
-# Select location on map for starting point
+# Select location on map for starting point using cursor and click
 pt <- unlist(locator(n = 1))
@@
 Flowpath B encounters a large pond along the way, significantly increasing residence time and N transformation (Table 3).
-These two undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet.
-If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky "eastern region (flowpath B) underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing N transformation.
+These two currently undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet.
+If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky "eastern region (flowpath B) underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing opportunities for N transformation.
 
-```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations with differing N removal estimates.
-Hydrofeatures are pictured on the  Nitrongen Transport Index map."}
+```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations within the Niantic River HUC-12 with differing N removal estimates.
+Hydrofeatures are pictured on the Nitrogen Transport Index map."}
 
@@
                      "#F9C159", "#F99844", "#F56A2E","#EF2820"))
-plot(niantic_data$streams, add = TRUE, col = "dodgerblue")
+huc <- niantic_data$huc
+streams <- st_intersection(huc, niantic_data$streams)
+plot(streams, add = TRUE, col = "dodgerblue")
 plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
@@
 
-```{r compare_flowpaths, include = TRUE, echo = FALSE}
+```{r compare-flowpaths, include = TRUE, echo = FALSE}
 knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B")
@@
 Variances granted on a case-by-case basis can quickly undermine local water quality goals.
-Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize acquisition or wetland restoration projects.
+Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize land acquisition or wetland restoration projects.
 
@@
 
-```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest, highlighting two developed areas."}
+```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two developed areas."}
 box <- c(1949902, 1953930, 2288202, 2292262)
@@
 ```
-```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index over land cover, with hotter colors denoting higher N delivery as estimated based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
+```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index over land cover, with hotter colors denoting higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
 case2_delivery <- crop(niantic_static_maps$delivery_idx, e)

### 2023-04-26 manuscript/manuscript.Rmd
Add phrase re: off-network hydrology
Add processed data needed for analysis to testdata.rda. Change eval = FALSE for downloading and prepping data.

@@
 #load(system.file("testdata.rda", package = "nsink"))
-#load("testdata.rda")
+load("testdata.rda")
 ```
@@
 
-```{r get-data, eval=TRUE, cache=FALSE, include=FALSE}
+```{r get-data, eval=FALSE, cache=FALSE, include=FALSE}
 # eval = TRUE if you want to download the data
@@
 
-```{r prep-data, eval = TRUE, include = FALSE}
+```{r prep-data, eval = FALSE, include = FALSE}
 # EPSG for CONUS Albers Equal Area
@@
 library(sf)
+# EPSG for CONUS Albers Equal Area
+aea <- 5072
+
 ## Linking to GEOS 3.8.1, GDAL 3.1.4, PROJ 6.3.1
@@
 None had off network stream or canal segments, while six of the seven had off-network waterbodies.
-These off-network waterbodies represented from 2 to 13% of total lake area.
+These off-network waterbodies represented from 2 to 14% (mean of 7%) of total lake area.
 We estimated removal from off-network stream segments to be equal to the median removal of all first order streams in the same catchment.

### 2023-04-26 manuscript/manuscript.Rmd
edits on text

@@
 output:
-  html_document:
-    df_print: paged
   word_document:
@@
 ## Download the data
+The first step for an N-Sink analysis with the `nsink` package is to download the required datasets.
+The only required information to download the data is a HUC identifier.
+The `nsink` package was developed using 12-digit HUC IDs, but larger HUCs (e.g.
+8-digit) may also be used.
+There are two functions provided for downloading the data.
+
 `nsink_get_huc_id()` - search 12-digit HUC IDs using a known location name.
 
-`nsink_get_data()` - download and save in a data folder hydrography, soils, and land cover for the specified 12-digit HUC.
+`nsink_get_data()` - download and save, in a specified folder, hydrography, soils, and land cover data for the specified 12-digit HUC.
 These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC.
@@
 Once that flowpath intersects the surface water network, flow is determined by flow along the NHD Plus surface water network.
-With a flowpath generated the cumulative N removal along that flowpath can be calculated with the `nsink_summarize_flowpath()` function, taking the flowpath and removal as input.
+With a flowpath generated, the cumulative N removal along that flowpath can be calculated with the `nsink_summarize_flowpath()` function, taking the flowpath and removal as input.
 A data frame is returned with each segment identified by type of sink, if present (i.e., "Hydric", "Stream", "Lake/Pond", or "No Removal"), the percent removal associated with that segment, and cumulative removal.
-Total cumulative removal is 100 - minimum of the `n_out` column.
+Total cumulative removal is 100 - the minimum of the `n_out` column.
 That is, `n_out` keeps track of the percent of N leaving each subsequent downstream segment (Table 2).
@@
 - `removal_effic`: HUC-wide estimate of potential nitrogen removal from different landscape sinks as a percentage.
-- `loading_idx`: An normalized index of nitrogen loads by land cover class derived from published sources, ranging from 0 to 1.
+- `loading_idx`: A normalized index of nitrogen loads by land cover class derived from published sources, ranging from 0 to 1.
 - `transport_idx`: N transport for a sample of all possible flowpaths in a given HUC.
@@
 [-@mayer2007meta] as having 80% N removal.
-In the current version N removal is cumulative as a flowpath intercepts hydric soils in any cell before encountering surface water.
+In the current version, N removal is cumulative as a flowpath intercepts hydric soils in any cell before encountering surface water.
 
@@
 
+SHOULD WE ADD SOMETHING ABOUT AREAS WHERE GROUNDWATER FLOW DOMINATES (E.G. CAPE COD)?
+
 Use Examples
@@
 niantic_removal <- nsink_calc_removal(niantic_data)
-niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal) # table of removal within flowpath segments
+# table of removal within flowpath segments
+niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal) 
 ```

### 2023-04-27 manuscript/manuscript.Rmd
Updated rmd with working version of nsink

@@
 niantic_download <- nsink_get_data(niantic_huc_id, 
-                                  data_dir = "nsink_niantic_data")
+                                  data_dir = "nsink_niantic_data/")
 ```

### 2023-04-27 manuscript/manuscript.Rmd
no longer loading rda

@@
 #load(system.file("testdata.rda", package = "nsink"))
-load("testdata.rda")
+#load("testdata.rda")
 ```

### 2023-04-27 manuscript/manuscript.Rmd
eval true

@@
 
-```{r get-data, eval=FALSE, cache=FALSE, include=FALSE}
+```{r get-data, eval=TRUE, cache=FALSE, include=FALSE}
 # eval = TRUE if you want to download the data

### 2023-04-28 R/example_ggplot_fig4.R
example ggplot for figure 4

@@ -1,0 +1,54 @@
+library(ggplot2)
+library(sf)
+library(terra)
+library(tidyterra)
+library(nsink)
+library(dplyr)
+
+
+aea <- 5072
+# Select location on map for starting point using cursor and click
+pta <- c(1954039, 2291592)
+start_loc_inter_a <- st_sf(st_sfc(st_point(pta), crs = aea))
+niantic_fp_a <- nsink_generate_flowpath(start_loc_inter_a, niantic_data)
+niantic_fp_removal_a <- nsink_summarize_flowpath(niantic_fp_a, niantic_removal)
+ptb <- c(1953529, 2290811)
+niantic_huc_id <- nsink_get_huc_id("Niantic River")$huc_12
+niantic_data <- nsink_prep_data(niantic_huc_id, projection = aea,
+                                data_dir = "nsink_niantic_data")
+niantic_removal <- nsink_calc_removal(niantic_data)
+start_loc_inter_b <- st_sf(st_sfc(st_point(ptb), crs = aea))
+niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 450)
+
+niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
+niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)
+huc <- niantic_data$huc
+streams <- st_intersection(huc, niantic_data$streams)
+x <- c(pta[1], ptb[1])
+y <- c(pta[2], ptb[2])
+labels <- c("A", "B")
+pts_df <- data.frame(x,y,labels)
+ab_plot <- st_as_sf(pts_df, coords = c("x","y"), crs = aea)
+
+# ggplot stuff starts here
+
+from <- c(seq(0,90,10), NA)
+to <- c(seq(10,100,10), NA)
+matrix(c(from,to,to),ncol = 3)
+tnsport_idx <- classify(rast(niantic_static_maps$transport_idx),matrix(c(from,to,to),ncol = 3))
+values(tnsport_idx) <- factor(values(tnsport_idx))
+ggplot() +
+  geom_spatraster(data = tnsport_idx) +
+  scale_fill_manual(values = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", 
+                                        "#E2F088", "#F5E871", "#F9C159", 
+                                        "#F99844", "#F56A2E","#EF2820"), 
[…]

### 2023-05-12 manuscript/manuscript.Rmd
example ggplot for figure 4
Incorporate Ken's edits

@@
 csl: ecology.csl
-abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
+abstract: Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
 Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery.
-Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system.
-Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to understand the effects of their decisions on downstream resources.
+Natural features within watersheds can act as landscape sinks for N, such as wetlands, streams and ponds that transform dissolved N into gaseous N, effectively removing it from the aquatic system.
+Identifying and evaluating these landscape sinks and their spatial relationship to N sources can help managers understand the effects of alternative decisions on downstream resources.
 N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
 A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index.
-To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England.
-The R package can be used to develop N delivery maps for a HUC-12 of interest, or users can visit the University of Connecticut website to explore the already mapped areas.
+To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and applied to produce static maps for all HUC-12 catchments in southern New England.
+The R package can be used to develop N delivery maps for a specific HUC-12 of interest, or users can visit the University of Connecticut website to explore the already mapped areas.
 Users can also investigate specific flowpaths interactively by clicking on any location within the catchment.
@@
 Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations.
-The intent of N-Sink, and its associated R package `nsink`, is as a screening tool to identify source areas where downgradient N sinks are able to process and remove excess N, versus areas where downgradient N sinks are less prevalent or effective at removing N.
-The latter argues for more aggressive N management at the source, while the former argues for protecting and/or restoring downgradient N sinks that are helping to reduce N delivery to estuaries.
+N-Sink, and its associated R package `nsink`, were developed to be a screening tool that describes the areas or points within a watershed linked with flow paths where downgradient N sinks process and remove excess N, versus areas where downgradient N sinks are less prevalent or less effective at removing N.
+This watershed flow path approach provides a means to easily visualize the relative sensitivity of an area within a watershed that may require more aggressive N management at the source, protection of existing N retentive areas, or restoration and construction of downgradient N retentive areas that are needed to reduce N delivery to estuaries.
 
@@
 [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
-Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
+Residence time within landscape features that support N removal is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
 
 The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] .
-Datasets used by `nsink` are the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web].
-The static maps include 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed.
+Datasets used by `nsink` include the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web].
+The static maps illustrate 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed.
 This heat map highlights “leaky” source areas with less downstream N retention versus those with higher downstream retention; and c) N Delivery Index, combining maps (1) and (2) along with source loading based on NLCD categories and expressed as an index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream.
+N-Sink and `nsink` integrate these data to make the application of complex hydrologic and biogeochemical processes accessible to end users and stakeholders in a visually clear and useful format.
+Here we describe the development of the package, and example workflow, and simple use applications to demonstrate the capabilities of the tools.
 
@@
 The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder.
-That version required time-consuming data manipulation by hand due to limitations of earlier NHDPlus datasets as well as the limitations of working in a GIS environment requiring a user license.
+The initial version required time-consuming data manipulation by hand due to limitations of earlier NHDPlus datasets as well as the limitations of working in a GIS environment requiring a user license.
 With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink was automated and applied to other HUC-12 catchments, leading to the development of the `nsink` package.
@@
 These static maps can be quickly plotted with `nsink_plot()` using the `map` argument to identify map type.
+These plots are designed for exploration and not necessarily for high quality maps.
+For higher quality maps, the source data can be plotted in software of a users choice (e.g.
+QGIS, `ggplot2`).
 

### 2023-06-05 manuscript/manuscript.Rmd
Change Fig 4 coding to use ggplot() and spatraster

@@
 streams <- st_intersection(huc, niantic_data$streams)
-plot(streams, add = TRUE, col = "dodgerblue")
-plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
-plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray40", lwd = 3.0)
-plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray40", lwd = 3.0)
-plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
-plot(st_geometry(niantic_fp_b$flowpath_ends), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
 x <- c(pta[1], ptb[1])
@@
 text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)
+from <- c(seq(0,90,10), NA)
+to <- c(seq(10,100,10), NA)
+#matrix(c(from,to,to),ncol = 3)
+tnsport_idx <- classify(rast(niantic_static_maps$transport_idx),matrix(c(from,to,to),ncol = 3))
+values(tnsport_idx) <- factor(values(tnsport_idx))
+ggplot() +
+  geom_spatraster(data = tnsport_idx) +
+  scale_fill_manual(values = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", 
+                                        "#E2F088", "#F5E871", "#F9C159", 
+                                        "#F99844", "#F56A2E","#EF2820"), 
+                                        na.value = NA, na.translate = FALSE) +
+  geom_sf(data = streams, color = "dodgerblue", linewidth = 0.7) +
+  geom_sf(data = niantic_data$lakes, fill = "dodgerblue", color = "dodgerblue") +
+  geom_sf(data = niantic_fp_a$flowpath_network, color = "grey40", linewidth = 2.5) +
+  geom_sf(data = niantic_fp_a$flowpath_ends, color = "grey40", linewidth = 2.5) +
+  geom_sf(data = niantic_fp_b$flowpath_network, color = "grey0", linewidth = 0.9) +
+  geom_sf(data = niantic_fp_b$flowpath_ends, color = "grey0", linewidth = 0.9) +
+  geom_sf(data = ab_plot, pch = 16, size = 2) + 
+  geom_sf_label(data = ab_plot, aes(label = labels))
 ```

### 2023-06-07 manuscript/manuscript.Rmd
Add packages for figs 4 and 5. Add new code for fig 5 and table 3, which accompanies fig 5.

--- 2023-05-12 manuscript/manuscript.Rmd
@@
 ---
-title: Assessing landscape N removal in coastal New England catchments using the N-Sink
-  approach with a new R package, nsink
+title: "Assessing landscape N removal in coastal New England catchments using the
+  N-Sink approach with a new R package, nsink"
 author:
@@
     reference_docx: manuscript_template.docx
+  pdf_document: default
+  html_document:
+    df_print: paged
 editor_options:
@@
 csl: ecology.csl
-abstract: Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
-Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery.
-Natural features within watersheds can act as landscape sinks for N, such as wetlands, streams and ponds that transform dissolved N into gaseous N, effectively removing it from the aquatic system.
-Identifying and evaluating these landscape sinks and their spatial relationship to N sources can help managers understand the effects of alternative decisions on downstream resources.
-N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
-A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index.
-To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and applied to produce static maps for all HUC-12 catchments in southern New England.
-The R package can be used to develop N delivery maps for a specific HUC-12 of interest, or users can visit the University of Connecticut website to explore the already mapped areas.
-Users can also investigate specific flowpaths interactively by clicking on any location within the catchment.
-A flowpath is generated, along with a table describing N removal along each segment.
+abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
+  quality. Land management decisions made within estuarine watersheds have a direct
+  impact on downstream N delivery. Natural features within watersheds can act as landscape
+  sinks for N, such as wetlands, streams and ponds that transform dissolved N into
+  gaseous N, effectively removing it from the aquatic system. Identifying and evaluating
+  these landscape sinks and their spatial relationship to N sources can help managers
+  understand the effects of alternative decisions on downstream resources. N-Sink
+  is an approach that uses widely available GIS data to identify landscape sinks within
+  HUC-12 (or larger) catchments, estimate their N removal potential and map the effect
+  of those sinks on N movement through the catchment. A series of static maps are
+  produced to visualize N removal efficiency, transport and delivery, the latter in
+  the form of an index. To facilitate data acquisition, processing and visualization,
+  an R package `nsink` was developed and applied to produce static maps for all HUC-12
+  catchments in southern New England. The R package can be used to develop N delivery
+  maps for a specific HUC-12 of interest, or users can visit the University of Connecticut
+  website to explore the already mapped areas. Users can also investigate specific
+  flowpaths interactively by clicking on any location within the catchment. A flowpath
+  is generated, along with a table describing N removal along each segment."
 ---
@@
 library(raster)
+library(ggplot2)
+library(terra)
+library(tidyterra)
 ```
@@
 
-```{r prep-data, eval = FALSE, include = FALSE}
+```{r prep-data, eval = TRUE, include = FALSE}
 # EPSG for CONUS Albers Equal Area
@@
 ```{r case1, echo = TRUE, eval = FALSE}
-# EPSG for CONUS Albers Equal Area
-aea <- 5072
+
 # Select location on map for starting point using cursor and click
@@
 
+pta <- c(1954039, 2291592)
+start_loc_inter_a <- st_sf(st_sfc(st_point(pta), crs = aea))
+niantic_fp_a <- nsink_generate_flowpath(start_loc_inter_a, niantic_data)
+niantic_fp_removal_a <- nsink_summarize_flowpath(niantic_fp_a, niantic_removal)
+
 ptb <- c(1953529, 2290811)
 start_loc_inter_b <- st_sf(st_sfc(st_point(ptb), crs = aea))
-
 niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
 niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)
-raster::plot(niantic_static_maps$transport_idx, breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), 
-             col = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", "#E2F088", "#F5E871",
-                     "#F9C159", "#F99844", "#F56A2E","#EF2820"))
+
 huc <- niantic_data$huc
 streams <- st_intersection(huc, niantic_data$streams)
-plot(streams, add = TRUE, col = "dodgerblue")
-plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
-plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray40", lwd = 3.0)
-plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray40", lwd = 3.0)
-plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
-plot(st_geometry(niantic_fp_b$flowpath_ends), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
 x <- c(pta[1], ptb[1])
 y <- c(pta[2], ptb[2])
-ab <- st_multipoint(matrix(c(x,y), nrow = 2, ncol = 2))
-ab_plot <- st_sf(st_sfc(ab), crs = aea)
-plot(ab_plot, add = TRUE, pch = 16)
-text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)
+labels <- c("A", "B")
+pts_df <- data.frame(x,y,labels)
+ab_plot <- st_as_sf(pts_df, coords = c("x","y"), crs = aea)
+
+from <- c(seq(0,90,10), NA)
+to <- c(seq(10,100,10), NA)
+
+tnsport_idx <- classify(rast(niantic_static_maps$transport_idx),matrix(c(from,to,to),ncol = 3))
+values(tnsport_idx) <- factor(values(tnsport_idx))
+
+ggplot() +
+  geom_spatraster(data = tnsport_idx) +
+  scale_fill_manual(values = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", 
+                                        "#E2F088", "#F5E871", "#F9C159", 
+                                        "#F99844", "#F56A2E","#EF2820"), 
+                                        na.value = NA, na.translate = FALSE) +
+  geom_sf(data = streams, color = "dodgerblue", linewidth = 0.7) +
+  geom_sf(data = niantic_data$lakes, fill = "dodgerblue", color = "dodgerblue") +
+  geom_sf(data = niantic_fp_a$flowpath_network, color = "grey40", linewidth = 2.5) +
+  geom_sf(data = niantic_fp_a$flowpath_ends, color = "grey40", linewidth = 2.5) +
+  geom_sf(data = niantic_fp_b$flowpath_network, color = "grey0", linewidth = 0.9) +
+  geom_sf(data = niantic_fp_b$flowpath_ends, color = "grey0", linewidth = 0.9) +
+  geom_sf(data = ab_plot, pch = 16, size = 2) + 
+  geom_sf_label(data = ab_plot, aes(label = labels))
 ```
@@
 Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled.
-Figure 5B shows the N Delivery Index from these two developed areas.
-Note that the northern area is less densely developed than the southern area but the N Delivery Index is higher in the northern area as compared to the southern area, suggesting that N source controls within the northern development might be considered.
 
-```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two developed areas."}
-box <- c(1949902, 1953930, 2288202, 2292262)
-e <- extent(box)
+```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (Table 3)."}
+
+box <- c(1950000, 1953800, 2289700, 2293100)
+e <- ext(box)
 legend_nlcd <- pal_nlcd()
-case2 <- crop(niantic_data$nlcd, e)
-vals <- unique(case2@data@values)
-leg <- legend_nlcd[legend_nlcd$ID %in% vals,]
-case2_north <- st_sf(st_sfc(st_point(c(1950846, 2290912)), crs = aea)) 
-case2_south <- st_sf(st_sfc(st_point(c(1952693, 2289249)), crs = aea))
+case2 <- crop(rast(niantic_data$nlcd), e)
+vals <- unique(case2$Class)
+leg <- legend_nlcd[legend_nlcd$ID %in% vals$Class,]
+values(case2) <- factor(values(case2))
 
-my_key <- list(rectangles=list(col = leg$Color),
-              text=list(lab=leg$Class),
-              space='left',
-              columns=1,
-              size=2,
-              cex=.6)
-colors <- legend_nlcd$Color
-plot(case2, col = legend_nlcd$Color)
-plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
-plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
+bdist <- units::set_units(0.5, km)
+ptw <- st_point(x = c(1951000, 2290650))
+ptw <- ptw %>% st_sfc(crs = aea)
+ptw_buf <- st_buffer(ptw, bdist)
+
+pte <- st_point(x = c(1952900, 2291750))
+pte <- pte %>% st_sfc(crs = aea)
+pte_buf <- st_buffer(pte, bdist)
+
+ggplot() +  geom_spatraster(data = case2, aes(fill = Class)) +
+  scale_fill_manual(values = leg$Color, na.value = NA, na.translate = FALSE,
+                    labels = leg$Class) + labs(fill = "NLCD Class") +
+  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
+        legend.text = element_text(size = 7), 
+        legend.title = element_text(size = 10)) +
+  geom_sf(data = ptw_buf, color = "black", fill = NA, linewidth = 1.5) +
+  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5)
 ```
-```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index over land cover, with hotter colors denoting higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
-case2_delivery <- crop(niantic_static_maps$delivery_idx, e)
-plot(case2, col = legend_nlcd$Color, alpha = 0.6, legend = FALSE)
-plot(case2_delivery, breaks = c(20, 40, 60, 80, 100), col = c("#FFBEBE", "#F57A7A","#A80405","#652600"), add = TRUE)
-plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
-plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
 
+```{r table3-build, eval = TRUE, include = FALSE}
+table3 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
+                                      "Wetlands", "Other"),
+                     "West (%)" = c("23","7","64","6","0"),
+                     "East (%)" = c("20","0","61","18","1"),
+                     check.names = FALSE)
+```
+```{r table3, include = TRUE, echo = FALSE}
+knitr::kable(table3, caption = "Table 3: Summary of Land Cover in Highlighted Areas of Interest")
+```
+
+An analysis of N delivery from these two similarly developed areas, however, show the western area is leakier than the eastern area, as assessed by summing the delivery index values (Figure 5B; West = 15759, East = 6770), suggesting that N source controls within the western development might be considered.
+
+```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas.
+Hotter colors denote higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
+
+from <- c(seq(0,90,10), NA)
+to <- c(seq(10,100,10), NA)
+delivery_idx <- classify(rast(niantic_static_maps$delivery_idx),matrix(c(from,to,to),ncol = 3))
+case2_delivery <- crop(delivery_idx, e)
+case2_delivery[case2_delivery < 20] <- 10
+values(case2_delivery) <- factor(values(case2_delivery))
+
+cols10 <-  c("white", "#FFBEBE", "#FAA8A8", "#FA9191", "#F57A7A",
+           "#DB5252", "#C22B2C", "#A80405",
+           "#910F04", "#7B1B02", "#652600")
+
+ggplot() + geom_spatraster(data = case2_delivery) +
+  scale_fill_manual(values = cols10, na.value = "white", na.translate = FALSE) +
+  guides(fill = guide_legend(reverse = TRUE)) + 
+  labs(fill = "") +
+  geom_sf(data = ptw_buf, color = "black", fill = NA, linewidth = 1.5) +
+  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5)
 ```

### 2023-06-08 manuscript/manuscript.Rmd
Updates based on EPA reviews done.

@@
   maps for a specific HUC-12 of interest, or users can visit the University of Connecticut
-  website to explore the already mapped areas. Users can also investigate specific
+  website to explore previously mapped areas. Users can also investigate specific
   flowpaths interactively by clicking on any location within the catchment. A flowpath
@@
 Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations.
-N-Sink, and its associated R package `nsink`, were developed to be a screening tool that describes the areas or points within a watershed linked with flow paths where downgradient N sinks process and remove excess N, versus areas where downgradient N sinks are less prevalent or less effective at removing N.
+N-Sink, and its associated R package `nsink`, were developed to be a screening tool that describes the areas or points within a watershed linked with flow paths where downgradient N sinks process and remove excess N, versus areas where downgradient N sinks are less prevalent or less effective at removing N [@hollister2022nsink; @R].
 This watershed flow path approach provides a means to easily visualize the relative sensitivity of an area within a watershed that may require more aggressive N management at the source, protection of existing N retentive areas, or restoration and construction of downgradient N retentive areas that are needed to reduce N delivery to estuaries.
@@
 [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
-Residence time within landscape features that support N removal is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
+Residence time within landscape features that support N removal through denitrification is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
 
@@
 The static maps illustrate 1) N Removal Efficiency, a color-coded map of the estimated N removal efficiency of the different landscape sinks; 2) N Transport Index, a heat map with the cumulative N removal along flowpaths originating from sources represented by a grid of points across a watershed.
-This heat map highlights “leaky” source areas with less downstream N retention versus those with higher downstream retention; and c) N Delivery Index, combining maps (1) and (2) along with source loading based on NLCD categories and expressed as an index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, taking into account the potential N removal as water moves downstream.
+This heat map highlights “leaky” source areas with less downstream N retention versus those with higher downstream retention; and 3) N Delivery Index, combining maps (1) and (2) along with source loading based on NLCD categories and expressed as an index ranging from 0 to 1, resulting in a map showing potential N delivery from different sources, after accounting for the potential N removal as water moves downstream.
 N-Sink and `nsink` integrate these data to make the application of complex hydrologic and biogeochemical processes accessible to end users and stakeholders in a visually clear and useful format.
 Here we describe the development of the package, and example workflow, and simple use applications to demonstrate the capabilities of the tools.
+
+<!-- EPA Review Comment: If the paper is about N-Sink integrating nsink it feels like the introduction could benefit from some more mention of N-Sink’s value-add at the end of the intro.
+Even just a sentence about a pre-analyzed web map being accessible to people who aren’t familiar with R.-->
 
@@
 Additionally, a convenience function, `nsink_build()`, is included that will run all of the required functions for a specified HUC.
-The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Efficiency, and N Delivery Index.
+The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Index, and N Delivery Index.
 
@@
 The only required information to download the data is a HUC identifier.
-The `nsink` package was developed using 12-digit HUC IDs, but larger HUCs (e.g.
+The `nsink` package was developed using 12-digit HUC IDs from NHDPlus, but larger HUCs (e.g.
 8-digit) may also be used.
@@
 
-`nsink_get_huc_id()` - search 12-digit HUC IDs using a known location name.
+- `nsink_get_huc_id()` - search 12-digit HUC IDs using a known location name.
 
-`nsink_get_data()` - download and save, in a specified folder, hydrography, soils, and land cover data for the specified 12-digit HUC.
+- `nsink_get_data()` - download and save, in a specified folder, hydrography, soils, and land cover data for the specified 12-digit HUC.
 These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC.
@@
 - raster_method: Contains a raster-based approach to calculating removal.  Used for the static maps showing potential N removal within landscape sinks.
-- land_removal: Represents land-based nitrogen removal within vegetated hydric soils (i.e., with impervious surface/developed areas removed).
+- land_removal: Represents land-based nitrogen removal within vegetated hydric soils with impervious surface removed).
 - network_removal: Contains removal along the NHD Plus flow network. Removal is calculated separately for streams and waterbodies (i.e., lakes and reservoirs).
@@
 
-SHOULD WE ADD SOMETHING ABOUT AREAS WHERE GROUNDWATER FLOW DOMINATES (E.G. CAPE COD)?
+<!-- SHOULD WE ADD SOMETHING ABOUT AREAS WHERE GROUNDWATER FLOW DOMINATES (E.G. CAPE COD)? -->
 
@@
 
-The maps produced by N-Sink can be helpful in prioritizing source controls, sink protection and/or restoration.
+The maps produced by N-Sink can helpl prioritize source controls, sink protection and/or restoration.
 For example, many wetlands are considered "protected" by federal, state or local laws and regulations; but as occurs with many types of protected areas [@neal2018coastal] they are frequently altered through variances granted by local commissions or committees charged with zoning decisions.

### 2023-06-27 manuscript/manuscript.Rmd
typo
shift in order of document template. Not significant.

--- 2023-06-07 manuscript/manuscript.Rmd
@@
 output:
+  html_document:
+    df_print: paged
+  pdf_document: default
   word_document:
     reference_docx: manuscript_template.docx
-  pdf_document: default
-  html_document:
-    df_print: paged
 editor_options:

### 2023-06-27 manuscript/manuscript.Rmd

--- 2023-06-08 manuscript/manuscript.Rmd
@@
 
-The maps produced by N-Sink can helpl prioritize source controls, sink protection and/or restoration.
+The maps produced by N-Sink can help prioritize source controls, sink protection and/or restoration.
 For example, many wetlands are considered "protected" by federal, state or local laws and regulations; but as occurs with many types of protected areas [@neal2018coastal] they are frequently altered through variances granted by local commissions or committees charged with zoning decisions.

### 2023-06-30 manuscript/manuscript.Rmd
Incorporates Art's and Chet's suggested edits. Most changes are in Use Case 2.

@@
 author:
-- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
+- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C. L.^*3*^, Gold, A. J. ^*1*^,
   Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
@@
   quality. Land management decisions made within estuarine watersheds have a direct
-  impact on downstream N delivery. Natural features within watersheds can act as landscape
-  sinks for N, such as wetlands, streams and ponds that transform dissolved N into
-  gaseous N, effectively removing it from the aquatic system. Identifying and evaluating
-  these landscape sinks and their spatial relationship to N sources can help managers
-  understand the effects of alternative decisions on downstream resources. N-Sink
-  is an approach that uses widely available GIS data to identify landscape sinks within
+  impact on downstream N delivery. Natural features frequently found within the 
+  headwaters of watersheds can act as landscape sinks for N, such as wetlands, streams 
+  and ponds that transform dissolved N into gaseous N, effectively removing it from 
+  the aquatic system. Identifying and evaluating these landscape sinks and their spatial
+  relationship to N sources can help managers understand the effects of alternative 
+  decisions on downstream resources. N-Sink is an approach that uses widely available 
+  GIS data to identify landscape sinks within
   HUC-12 (or larger) catchments, estimate their N removal potential and map the effect
@@
 Hydrologic connections and flowpaths are influential in the delivery of nutrients to estuaries [@mengistu2020wetland].
-Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations.
+Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale could provide watershed managers, land use planners and conservation organizations additional strategies that target N reductions.
 N-Sink, and its associated R package `nsink`, were developed to be a screening tool that describes the areas or points within a watershed linked with flow paths where downgradient N sinks process and remove excess N, versus areas where downgradient N sinks are less prevalent or less effective at removing N [@hollister2022nsink; @R].
@@
 [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds.
-Residence time within landscape features that support N removal through denitrification is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
+The approach relies on residence time within landscape features that support N removal as the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].
 
@@
 
-pta <- c(1954039, 2291592)
+#pta <- c(1954039, 2291592)
+pta <- c(1953529, 2290811)
 start_loc_inter_a <- st_sf(st_sfc(st_point(pta), crs = aea))
@@
 N-Sink was designed for catchments where surface and groundwater flow paths are not highly manipulated.
-In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage bypasses the processes within wetland ecosystems, the removal of N in riparian zones is likely to be overstated.
+In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage are present, flow can bypass the removal processes within wetland ecosystems [@gold2001landscape].
+In these situations first-order streams may be channelized, leading to higher velocities than in undisturbed channels and the N-Sink estimates of removal may be overstated.N-Sink does not account for nitrogen removal from high permeability (e.g., sand and gravel) unconfined aquifers along the coastal margin.
+Here, nitrogen-enriched groundwater may enter estuaries directly as submarine discharge rather than via baseflow to stream networks [@nowicki2008groundwater], requiring site specific assessments to evaluate N removal in salt marshes or as upwelling into the seabed [@addy2005denitrification;@robinson2018groundwater].
+These situations have been estimated to account for approximately 6% of freshwater inputs to estuaries globally and in areas of New England [@burnett2003groundwater;@nowicki2008groundwater].
+
 The variation in loading associated with different agricultural practices, density of unsewered development and inputs from wastewater treatment plants are not encompassed in the default loading index.
 Thus the numeric outputs of the tool focus on metrics suitable for planning purposes, such as percent removal of N from source to receiving water and relative intensity of loading from different locations.
-It is not intended to replace high-resolution models that rely on long term weather records and incorporate detailed spatial data on N loading, within-field transport or variable source hydrology (e.g., Easton [-@easton2008re]).
+It is not intended to replace high-resolution models that rely on long term weather records and incorporate detailed spatial data on N loading, within-field transport or variable source hydrology (e.g., Easton [-@easton2008re])).
 Instead, N-Sink combines widely available datasets for waterway networks, soils and land cover with nitrogen dynamics developed for these datasets to highlight major sources and sinks of nitrogen within a watershed context.
-
-<!-- SHOULD WE ADD SOMETHING ABOUT AREAS WHERE GROUNDWATER FLOW DOMINATES (E.G. CAPE COD)? -->
 
@@
 
-ptb <- c(1953529, 2290811)
+#ptb <- c(1953529, 2290811)
+ptb <- c(1954039, 2291592)
 start_loc_inter_b <- st_sf(st_sfc(st_point(ptb), crs = aea))
@@
 y <- c(pta[2], ptb[2])
-labels <- c("A", "B")
+labels <- c("B", "A")
 pts_df <- data.frame(x,y,labels)
@@
 
-An analysis of N delivery from these two similarly developed areas, however, show the western area is leakier than the eastern area, as assessed by summing the delivery index values (Figure 5B; West = 15759, East = 6770), suggesting that N source controls within the western development might be considered.
+```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Transport Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas."}
 
-```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas.
-Hotter colors denote higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
+from <- c(seq(0,90,10), NA)
+to <- c(seq(10,100,10), NA)
+
+tnsport_idx <- classify(rast(niantic_static_maps$transport_idx),matrix(c(from,to,to),ncol = 3))
+case2_tnsport <- crop(tnsport_idx, e)
+values(case2_tnsport) <- factor(values(case2_tnsport))
+ggplot() +
+  geom_spatraster(data = case2_tnsport) +
+  scale_fill_manual(values = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", 
+                               "#E2F088", "#F5E871", "#F9C159", 
+                               "#F99844", "#F56A2E","#EF2820"), 
+                    na.value = NA, na.translate = FALSE) +
+  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
+        legend.text = element_text(size = 7), 
+        legend.title = element_text(size = 10)) +
+  guides(fill = guide_legend(reverse = TRUE)) + 
+  labs(fill = "") +
+  geom_sf(data = ptw_buf, color = "black", fill = NA, linewidth = 1.5) +
+  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5)
+```
+
+An analysis of N delivery from these two similarly developed areas (average N loading index of developed areas: West = 0.57; East = 0.56),  show the western area may contribute more N loading to the estuary than the eastern area, with the average delivery index value of the western area (41) almost 80% higher than the eastern area (23; Figure 5c), suggesting a greater need for N source controls within the western development.
+
+```{r, plot5c, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5C: Nitrogen Delivery Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas.
+Darker colors denote higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency.
+"}
 

### 2023-07-03 manuscript/manuscript.Rmd
some improvement on figures.  Needs some more (especially with NLCD key)

@@
 output:
-  html_document:
-    df_print: paged
-  pdf_document: default
   word_document:
@@
   tidy = FALSE,
-  message = FALSE
+  message = FALSE,
+  fig.height = 4.8,
+  fig.width = 6
 )
@@
 
-These static maps can be quickly plotted with `nsink_plot()` using the `map` argument to identify map type.
+These static maps can be quickly plotted with the convenience function `nsink_plot()` using the `map` argument to identify map type.
 These plots are designed for exploration and not necessarily for high quality maps.
@@
 ```{r case1, echo = TRUE, eval = FALSE}
-
+# Plot watershed with base R
+plot(st_geometry(niantic_data$huc))
 # Select location on map for starting point using cursor and click
@@
 niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
+# Plot flowpath
+plot(st_geometry(niantic_fp$flowpath_network), add = TRUE)
 # Summarize N removal along the flowpath
@@
 ```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations within the Niantic River HUC-12 with differing N removal estimates.
-Hydrofeatures are pictured on the Nitrogen Transport Index map."}
+Hydrofeatures are pictured on the Nitrogen Transport Index map.", fig.height = 10, fig.width = 7.5}
 
@@
 
-ggplot() +
+map <- ggplot() +
   geom_spatraster(data = tnsport_idx) +
@@
   geom_sf(data = ab_plot, pch = 16, size = 2) + 
-  geom_sf_label(data = ab_plot, aes(label = labels))
+  geom_sf_label(data = ab_plot, aes(label = labels)) +
+  theme_bw() +
+  labs(x = "", y = "")
+map
+ggsave("fig4.jpg", map, height = 10, width = 7.5)
 ```
@@
 
-ggplot() +  geom_spatraster(data = case2, aes(fill = Class)) +
+map <- ggplot() +  geom_spatraster(data = case2, aes(fill = Class)) +
   scale_fill_manual(values = leg$Color, na.value = NA, na.translate = FALSE,
@@
   geom_sf(data = ptw_buf, color = "black", fill = NA, linewidth = 1.5) +
-  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5)
+  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5) +
+  theme_bw() +
+  labs(x = "", y = "")
+map
+ggsave("fig5a.jpg", map, height = 6.5, width = 7.5)
 ```
@@
 values(case2_tnsport) <- factor(values(case2_tnsport))
-ggplot() +
+map <- ggplot() +
   geom_spatraster(data = case2_tnsport) +
@@
   geom_sf(data = ptw_buf, color = "black", fill = NA, linewidth = 1.5) +
-  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5)
+  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5) +
+  theme_bw() +
+  labs(x = "", y = "")
+map
+ggsave("fig5b.jpg", map, height = 6, width = 7.5)
 ```
@@
 
-ggplot() + geom_spatraster(data = case2_delivery) +
+map <- ggplot() + geom_spatraster(data = case2_delivery) +
   scale_fill_manual(values = cols10, na.value = "white", na.translate = FALSE) +
@@
   geom_sf(data = ptw_buf, color = "black", fill = NA, linewidth = 1.5) +
-  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5)
+  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5) +
+  theme_bw() +
+  labs(x = "", y = "")
+map
+ggsave("fig5c.jpg", map, height = 6, width = 7.5)
 ```

### 2023-09-05 manuscript/manuscript.Rmd
suzy edits

@@
 - raster_method: Contains a raster-based approach to calculating removal.  Used for the static maps showing potential N removal within landscape sinks.
-- land_removal: Represents land-based nitrogen removal within vegetated hydric soils with impervious surface removed).
+- land_removal: Represents land-based nitrogen removal within vegetated hydric soils with impervious surface removed.
 - network_removal: Contains removal along the NHD Plus flow network. Removal is calculated separately for streams and waterbodies (i.e., lakes and reservoirs).
@@
 These static maps can be quickly plotted with the convenience function `nsink_plot()` using the `map` argument to identify map type.
+For example, `nsink_plot(niantic_static_maps, map = "removal")` will create a plot of Nitrogen Removal Efficiency (Figure 1).
+The other `map` argument options are "transport" and "delivery" (Figures 2-3).
 These plots are designed for exploration and not necessarily for high quality maps.
@@
 
-```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 1) for the Niantic River HUC-12, showing the estimated percentage of N removed within three types of landscape N sinks - wetlands, streams, lakes/ponds/reservoirs."}
+```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 1) for the Niantic River HUC-12, showing the estimated percentage of N removed within three types of landscape N sinks - wetlands, streams, lakes/ponds/reservoirs.", echo=FALSE}
 
@@
 ```
-```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (range 0 to 100%) for the Niantic River HUC-12, showing the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment."}
+```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (range 0 to 100%) for the Niantic River HUC-12, showing the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment.", echo=FALSE}
 
@@
 N loading estimates are based on land cover classes, making use of peer-reviewed published data, and normalized, with a range of 0 to 1.
-These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the NDI."}
+These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the NDI.", echo=FALSE}
 
@@
 In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage are present, flow can bypass the removal processes within wetland ecosystems [@gold2001landscape].
-In these situations first-order streams may be channelized, leading to higher velocities than in undisturbed channels and the N-Sink estimates of removal may be overstated.N-Sink does not account for nitrogen removal from high permeability (e.g., sand and gravel) unconfined aquifers along the coastal margin.
+In these situations first-order streams may be channelized, leading to higher velocities than in undisturbed channels and the N-Sink estimates of removal may be overstated.
+N-Sink does not account for nitrogen removal from high permeability (e.g., sand and gravel) unconfined aquifers along the coastal margin.
 Here, nitrogen-enriched groundwater may enter estuaries directly as submarine discharge rather than via baseflow to stream networks [@nowicki2008groundwater], requiring site specific assessments to evaluate N removal in salt marshes or as upwelling into the seabed [@addy2005denitrification;@robinson2018groundwater].
@@
 
-Users can use the following to interactively identify a starting point for a flowpath of interest. 
+Users can use the following code to interactively identify a starting point for a flowpath of interest. 
 ```{r case1, echo = TRUE, eval = FALSE}

### 2023-09-06 manuscript/manuscript.Rmd
Add a couple of sentences about use in LISS fertilizer project.

@@
 N-Sink is a decision support tool and was created to be a useful, easy way for local land use managers to explore the relationship between land use and nitrogen pollution in their waters.
+Currently N-Sink is being used as an integral component in a project based in Long Island Sound, funded by the [Long Island Sound Study Research Fund](https://longislandsoundstudy.net/2022-research-project-descriptions/).
+The project is focused on targeting outreach on fertilizer use within the 43,560 km2 catchment.
 Local decision-makers that have prioritized nitrogen mitigation in their long-term planning efforts can use this tool to help them better understand the potential impact of proposed development projects and zoning variances.

### 2023-09-07 manuscript/manuscript.Rmd
submission ready!  cache on, sessioninfo, akncowledgements, and code statement.

@@
   eval = TRUE,
-  cache = FALSE,
+  cache = TRUE,
   tidy = FALSE,
@@
 Here we describe the development of the package, and example workflow, and simple use applications to demonstrate the capabilities of the tools.
+The code for this analysis is available from <https://github.com/USEPA/nsink_manuscript/> and is archived at <https://doi.org/10.5281/zenodo.8325360>.
+All analyses were conducted with R version 4.2.2 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2022; @here; @ggplot2; @sf1; @sf2; @knitr1; @knitr2; @knitr3; @feddata; @hollister2022nsink; @units; @raster; @terra; @tidyterra; @lubridate; @units].
 
@@
 
+```{r session_info, echo=FALSE}
+#Send session info to a file: 
+capture.output(sessionInfo(),file=here::here("sessioninfo.txt"))
+```
+
+Acknowledgements
+================
+
+Many people have contributed in various ways to the development of the N-Sink 
+concept. In particular, we would like to thank Peter August and Chris Damon of 
+the University of Rhode Island’s Department of Natural Resources Science.  
+Additionally, we are grateful to Stephen Shivers, Anne Kuhn, Andrea Bartolotti, 
+Suzanne Ayvazian, and Tim Gleason for constructive reviews of this paper. The 
+views expressed in this article are those of the authors and do not necessarily 
+represent the views or policies of the U.S. Environmental Protection Agency. Any 
+mention of trade names, products, or services does not imply an endorsement by 
+the U.S. Government or the U.S. Environmental Protection Agency. The EPA does 
+not endorse any commercial products, services, or enterprises. This contribution 
+is identified by the tracking number ORD-054342 of the Atlantic Coastal 
+Environmental Sciences Division, Office of Research and Development, Center for 
+Environmental Measurement and Modeling, US Environmental Protection Agency.
+
 Bibliography

### 2023-09-29 manuscript/manuscript.Rmd
blank commit
Figure 4 fix. A and B are reversed.

--- 2023-09-06 manuscript/manuscript.Rmd
@@
 y <- c(pta[2], ptb[2])
-labels <- c("B", "A")
+labels <- c("A", "B")
 pts_df <- data.frame(x,y,labels)

### 2023-10-04 manuscript/manuscript.Rmd
Add Rachel as co-author, fix fig 4 (A to B), add note about terra package, edits to acknowledgements,

--- 2023-09-07 manuscript/manuscript.Rmd
@@
 author:
-- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C. L.^*3*^, Gold, A. J. ^*1*^,
-  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
+- "Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\\*^ ^*2*^, Arnold, C. L.^*3*^, Gold, A.
+  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Lei-Parent,
+  Qian*^*3*^, Forshay, K. J.^*4*^"
 - ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
@@
 output:
+  pdf_document: default
   word_document:
     reference_docx: manuscript_template.docx
+  html_document:
+    df_print: paged
 editor_options:
@@
   quality. Land management decisions made within estuarine watersheds have a direct
-  impact on downstream N delivery. Natural features frequently found within the 
-  headwaters of watersheds can act as landscape sinks for N, such as wetlands, streams 
-  and ponds that transform dissolved N into gaseous N, effectively removing it from 
-  the aquatic system. Identifying and evaluating these landscape sinks and their spatial
-  relationship to N sources can help managers understand the effects of alternative 
-  decisions on downstream resources. N-Sink is an approach that uses widely available 
-  GIS data to identify landscape sinks within
-  HUC-12 (or larger) catchments, estimate their N removal potential and map the effect
-  of those sinks on N movement through the catchment. A series of static maps are
-  produced to visualize N removal efficiency, transport and delivery, the latter in
-  the form of an index. To facilitate data acquisition, processing and visualization,
-  an R package `nsink` was developed and applied to produce static maps for all HUC-12
-  catchments in southern New England. The R package can be used to develop N delivery
-  maps for a specific HUC-12 of interest, or users can visit the University of Connecticut
-  website to explore previously mapped areas. Users can also investigate specific
-  flowpaths interactively by clicking on any location within the catchment. A flowpath
-  is generated, along with a table describing N removal along each segment."
+  impact on downstream N delivery. Natural features frequently found within the headwaters
+  of watersheds can act as landscape sinks for N, such as wetlands, streams and ponds
+  that transform dissolved N into gaseous N, effectively removing it from the aquatic
+  system. Identifying and evaluating these landscape sinks and their spatial relationship
+  to N sources can help managers understand the effects of alternative decisions on
+  downstream resources. N-Sink is an approach that uses widely available GIS data
+  to identify landscape sinks within HUC-12 (or larger) catchments, estimate their
+  N removal potential and map the effect of those sinks on N movement through the
+  catchment. A series of static maps are produced to visualize N removal efficiency,
+  transport and delivery, the latter in the form of an index. To facilitate data acquisition,
+  processing and visualization, an R package `nsink` was developed and applied to
+  produce static maps for all HUC-12 catchments in southern New England. The R package
+  can be used to develop N delivery maps for a specific HUC-12 of interest, or users
+  can visit the University of Connecticut website to explore previously mapped areas.
+  Users can also investigate specific flowpaths interactively by clicking on any location
+  within the catchment. A flowpath is generated, along with a table describing N removal
+  along each segment."
 ---
@@
 
-```{r setup, echo=FALSE, warning = FALSE}
+```{r setup, echo=FALSE, eval = TRUE, warning = FALSE}
 library(FedData)
@@
 Specifically, `nsink` relies on packages `sf` and `raster`.
+The `raster` package in R will soon be superseded by the `terra` package, which will be integrated in future `nsink` package versions.
 The `nsink` package is available from https://github.com/usepa/nsink and is fully described in [@hollister2022nsink].
@@
 These plots are designed for exploration and not necessarily for high quality maps.
-For higher quality maps, the source data can be plotted in software of a users choice (e.g.
+For higher quality maps, the source data can be plotted in software of a user's choice (e.g.
 QGIS, `ggplot2`).
@@
 
-```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (Table 3)."}
+```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (Table 4)."}
 
@@
 
-```{r table3-build, eval = TRUE, include = FALSE}
+```{r table4-build, eval = TRUE, include = FALSE}
 table3 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
@@
 ```
-```{r table3, include = TRUE, echo = FALSE}
+```{r table4, include = TRUE, echo = FALSE}
 knitr::kable(table3, caption = "Table 3: Summary of Land Cover in Highlighted Areas of Interest")
@@
 Additionally, we are grateful to Stephen Shivers, Anne Kuhn, Andrea Bartolotti, 
-Suzanne Ayvazian, and Tim Gleason for constructive reviews of this paper. The 
-views expressed in this article are those of the authors and do not necessarily 
+Suzanne Ayvazian, and Tim Gleason for constructive reviews of this paper.
+
+The views expressed in this article are those of the authors and do not necessarily 
 represent the views or policies of the U.S. Environmental Protection Agency. Any 
@@
 the U.S. Government or the U.S. Environmental Protection Agency. The EPA does 
-not endorse any commercial products, services, or enterprises. This contribution 
-is identified by the tracking number ORD-054342 of the Atlantic Coastal 
-Environmental Sciences Division, Office of Research and Development, Center for 
-Environmental Measurement and Modeling, US Environmental Protection Agency.
+not endorse any commercial products, services, or enterprises. This research has 
+been supported, in part, by a cooperative agreement with the U.S. Environmental
+Protection Agency's Office of Research and Development, EPA Assistant Agreement
+Number CR83680001, "N-Sink tool and outreach to two watersheds in New England."
+This contribution is identified by the tracking number ORD-054342 of the 
+Atlantic Coastal Environmental Sciences Division, Office of Research and Development, 
+Center for Environmental Measurement and Modeling, US Environmental Protection Agency.
 

### 2023-10-09 manuscript/manuscript.Rmd
Add Rachel as co-author, small edits to print out figs at 300 dpi, fix small typos or omissions.

@@
 author:
-- "Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\\*^ ^*2*^, Arnold, C. L.^*3*^, Gold, A.
-  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Lei-Parent,
-  Qian*^*3*^, Forshay, K. J.^*4*^"
+- "Kellogg, D. Q.^\\*^^*1*^, Hollister. J. W. ^*2*^, Arnold, C. L.^*3*^, Gold, A.
+  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.^*3*^, Lei-Parent,
+  Qian^*3*^, Forshay, K. J.^*4*^"
 - ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
@@
   Characterization and Remediation Division, Ada, OK 74820
-- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
+- '^\*^corresponding author: <qkellogg@uri.edu>'
 bibliography: manuscript.bib
 output:
-  pdf_document: default
   word_document:
+    keep_md: true
     reference_docx: manuscript_template.docx
@@
     df_print: paged
+  pdf_document: default
 editor_options:
@@
   fig.height = 4.8,
-  fig.width = 6
+  fig.width = 6,
+  dpi = 300
 )
@@
 
-```{r get-data, eval=TRUE, cache=FALSE, include=FALSE}
+```{r get-data, eval = TRUE, cache=FALSE, include=FALSE}
 # eval = TRUE if you want to download the data
 # Get data for selected HUC
-niantic_download <- nsink_get_data(niantic_huc_id, 
-                                  data_dir = "nsink_niantic_data/")
+nsink::nsink_get_data(niantic_huc_id, data_dir = "nsink_niantic_data/")
 ```
@@
 ```{r compare-flowpaths, include = TRUE, echo = FALSE}
-knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B")
+knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B.")
 ```
@@
 ```{r table4, include = TRUE, echo = FALSE}
-knitr::kable(table3, caption = "Table 3: Summary of Land Cover in Highlighted Areas of Interest")
+knitr::kable(table3, caption = "Table 4: Summary of land cover in highlighted areas of interest.")
 ```
@@
 
-An analysis of N delivery from these two similarly developed areas (average N loading index of developed areas: West = 0.57; East = 0.56),  show the western area may contribute more N loading to the estuary than the eastern area, with the average delivery index value of the western area (41) almost 80% higher than the eastern area (23; Figure 5c), suggesting a greater need for N source controls within the western development.
+An analysis of N delivery from these two similarly developed areas (average N loading index of developed areas: West = 0.57; East = 0.56),  show the western area may contribute more N loading to the estuary than the eastern area, with the average delivery index value of the western area (41) almost 80% higher than the eastern area (23; Figure 5C), suggesting a greater need for N source controls within the western development.
 
-```{r, plot5c, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5C: Nitrogen Delivery Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas.
+```{r, plot5C, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5C: Nitrogen Delivery Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas.
 Darker colors denote higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency.

### 2023-10-10 manuscript/manuscript.Rmd
Edit Table titles

@@
 ```{r table1, include = TRUE, echo = FALSE}
-knitr::kable(table1, caption = "Table 1: Functions in the R package `nsink`")
+knitr::kable(table1, caption = "Table 1: Functions in the R package, `nsink`, with short descriptions.")
 ```
@@
 ```{r show-summary, include = TRUE, echo = FALSE}
-knitr::kable(niantic_fp_removal_a, caption = "Table 2: Summary of N removal along specified flowpath A (See Figure 4)")
+knitr::kable(niantic_fp_removal_a, caption = "Table 2: Summary of N removal along specified flowpath A (Fig. 4), as generated by the R package, nsink.")
 ```
@@
 ```{r compare-flowpaths, include = TRUE, echo = FALSE}
-knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B.")
+knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B (Fig. 4), as generated by the R package, nsink.")
 ```

### 2023-10-25 manuscript/manuscript.Rmd
Small changes in preparation for submission to f1000Research

@@
 title: "Assessing landscape N removal in coastal New England catchments using the
-  N-Sink approach with a new R package, nsink"
+  N-Sink approach with the R package, nsink"
 author:
 - "Kellogg, D. Q.^\\*^^*1*^, Hollister. J. W. ^*2*^, Arnold, C. L.^*3*^, Gold, A.
-  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.^*3*^, Lei-Parent,
+  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C. B.^*3*^, Dickson, D. W.^*3*^, Lei-Parent,
   Qian^*3*^, Forshay, K. J.^*4*^"
@@
   within the catchment. A flowpath is generated, along with a table describing N removal
-  along each segment."
+  along each segment.In this software tool article we describe the motivation behind developing 
+  nsink, discuss the implementation in R, and describe the use of nsink with two use case examples. 
+  nsink is available from <https://github.com/USEPA/nsink>."
 ---
@@
 
-N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data.
+N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]) and makes extensive use of widely available GIS data.
 The theoretical underpinnings are outlined in Kellogg et al.
@@
 In this case we are assuming these first order segments behave similarly to other first order stream reaches.
-A subset of these off-network stream segments are identified by the NHD dataset as ditches and canals.
+A subset of these off-network stream segments is identified by the NHD dataset as ditches and canals.
 These are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore N-Sink assigns removal as the lower quartile of removal from the highest order streams in the same catchment.
@@
 In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage are present, flow can bypass the removal processes within wetland ecosystems [@gold2001landscape].
-In these situations first-order streams may be channelized, leading to higher velocities than in undisturbed channels and the N-Sink estimates of removal may be overstated.
+In these situations, first-order streams may be channelized, leading to higher velocities than in undisturbed channels and the N-Sink estimates of removal may be overstated.
 N-Sink does not account for nitrogen removal from high permeability (e.g., sand and gravel) unconfined aquifers along the coastal margin.
@@
 These two currently undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet.
-If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky "eastern region (flowpath B) underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing opportunities for N transformation.
+If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky" eastern region (flowpath B) underlines the importance of wetlands and slow-moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing opportunities for N transformation.
 
@@
 Similarly, sources of N that are contributing relatively more N due to their location within the catchment can be highlighted using the `nsink` static maps.
-The N Delivery Index map combines N sources with N transport to highlight developed areas that are located on "leakier parts" of a catchment.
+The N Delivery Index map combines N sources with N transport to highlight developed areas that are located on "leakier" parts of a catchment.
 Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled.

### 2023-10-26 manuscript/manuscript.Rmd
local changes

--- 2023-10-09 manuscript/manuscript.Rmd
@@
 author:
-- "Kellogg, D. Q.^\\*^^*1*^, Hollister. J. W. ^*2*^, Arnold, C. L.^*3*^, Gold, A.
-  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.^*3*^, Lei-Parent,
-  Qian^*3*^, Forshay, K. J.^*4*^"
+- "Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\\*^ ^*2*^, Arnold, C. L.^*3*^, Gold, A.
+  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Lei-Parent,
+  Qian*^*3*^, Forshay, K. J.^*4*^"
 - ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
@@
   Characterization and Remediation Division, Ada, OK 74820
-- '^\*^corresponding author: <qkellogg@uri.edu>'
+- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
 bibliography: manuscript.bib
 output:
+  pdf_document: default
   word_document:
@@
     df_print: paged
-  pdf_document: default
 editor_options:
@@
   fig.height = 4.8,
-  fig.width = 6,
-  dpi = 300
+  fig.width = 6
 )
@@
 
-```{r get-data, eval = TRUE, cache=FALSE, include=FALSE}
+```{r get-data, eval=TRUE, cache=FALSE, include=FALSE}
 # eval = TRUE if you want to download the data
 # Get data for selected HUC
-nsink::nsink_get_data(niantic_huc_id, data_dir = "nsink_niantic_data/")
+niantic_download <- nsink::nsink_get_data(niantic_huc_id, 
+                                  data_dir = "nsink_niantic_data/")
 ```
@@
 ```{r compare-flowpaths, include = TRUE, echo = FALSE}
-knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B.")
+knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B")
 ```
@@
 ```{r table4, include = TRUE, echo = FALSE}
-knitr::kable(table3, caption = "Table 4: Summary of land cover in highlighted areas of interest.")
+knitr::kable(table3, caption = "Table 3: Summary of Land Cover in Highlighted Areas of Interest")
 ```
@@
 
-An analysis of N delivery from these two similarly developed areas (average N loading index of developed areas: West = 0.57; East = 0.56),  show the western area may contribute more N loading to the estuary than the eastern area, with the average delivery index value of the western area (41) almost 80% higher than the eastern area (23; Figure 5C), suggesting a greater need for N source controls within the western development.
+An analysis of N delivery from these two similarly developed areas (average N loading index of developed areas: West = 0.57; East = 0.56),  show the western area may contribute more N loading to the estuary than the eastern area, with the average delivery index value of the western area (41) almost 80% higher than the eastern area (23; Figure 5c), suggesting a greater need for N source controls within the western development.
 
-```{r, plot5C, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5C: Nitrogen Delivery Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas.
+```{r, plot5c, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5C: Nitrogen Delivery Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas.
 Darker colors denote higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency.

### 2023-10-31 manuscript/manuscript.Rmd
Small edits before submitting to f1000Research

--- 2023-10-25 manuscript/manuscript.Rmd
@@
 
-```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 1) for the Niantic River HUC-12, showing the estimated percentage of N removed within three types of landscape N sinks - wetlands, streams, lakes/ponds/reservoirs.", echo=FALSE}
+```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 1) for the Niantic River HUC-12, showing the estimated fraction of N removed within three types of landscape N sinks - wetlands, streams, lakes/ponds/reservoirs.", echo=FALSE}
 
@@
 N loading estimates are based on land cover classes, making use of peer-reviewed published data, and normalized, with a range of 0 to 1.
-These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the NDI.", echo=FALSE}
+These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the N Delivery Index.", echo=FALSE}
 
@@
 
-```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (Table 4)."}
+```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (summary given in Table 4)."}
 
@@
 ```{r table4-build, eval = TRUE, include = FALSE}
-table3 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
+table4 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
                                       "Wetlands", "Other"),
@@
 ```{r table4, include = TRUE, echo = FALSE}
-knitr::kable(table3, caption = "Table 4: Summary of land cover in highlighted areas of interest.")
+knitr::kable(table4, caption = "Table 4: Summary of land cover in highlighted areas of interest.")
 ```

### 2024-03-06 manuscript/manuscript.Rmd
Changes made to reflect revised version submitted to f1000Research. Also includes new code that calculates table 4 numbers during knit.

@@
 ---
-title: "Assessing landscape N removal in coastal New England catchments using the
-  N-Sink approach with the R package, nsink"
+title: "Assessing landscape N removal in coastal New England catchments using the N-Sink approach with the R package, *nsink*"
 author:
-- "Kellogg, D. Q.^\\*^^*1*^, Hollister. J. W. ^*2*^, Arnold, C. L.^*3*^, Gold, A.
-  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C. B.^*3*^, Dickson, D. W.^*3*^, Lei-Parent,
-  Qian^*3*^, Forshay, K. J.^*4*^"
-- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
-  RI 02881
-- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
-  Coastal Environmental Sciences Division, Narragansett, RI 02882
-- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
-  CT 06438
-- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
-  Characterization and Remediation Division, Ada, OK 74820
-- '^\*^corresponding author: <qkellogg@uri.edu>'
+- "Dorothy Q Kellogg^*1*^^\\*^, Jeffrey W.
+Hollister^*2*^, Chester L.
+Arnold^*3*^, Arthur J.
+Gold^*1*^, Emily H.
+Wilson^*3*^, Cary B.
+Chadwick^*3*^, David W.
+Dickson^*3*^, Qian Lei-Parent^*3*^, Kenneth J.
+Forshay^*4*^"
+- ^*1*^ Department of Natural Resources Science, University of Rhode Island, Kingston, RI 02881, USA
+- ^*2*^ Office Of Research and Development, Atlantic Coastal Environmental Sciences Division, US Environmental Protection Agency, Narragansett, RI 02882, USA
+- ^*3*^ Center for Land Use Education and Research, University of Connecticut, Haddam, CT 06438, USA
+- ^*4*^ Office Of Research and Development, Groundwater Characterization and Remediation Division, US Environmental Protection Agency, Ada, OK 74820, USA
+- '^\*^Corresponding Author: 
+Dorothy Q Kellogg
+Department of Natural Resources Science
+University of Rhode Island, Kingston, RI 02881, USA
+Email address: <qkellogg@uri.edu>'
 bibliography: manuscript.bib
@@
     keep_md: true
-    reference_docx: manuscript_template.docx
+    reference_docx: manuscript_f1000_template.docx
   html_document:
@@
 csl: ecology.csl
-abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
-  quality. Land management decisions made within estuarine watersheds have a direct
-  impact on downstream N delivery. Natural features frequently found within the headwaters
-  of watersheds can act as landscape sinks for N, such as wetlands, streams and ponds
-  that transform dissolved N into gaseous N, effectively removing it from the aquatic
-  system. Identifying and evaluating these landscape sinks and their spatial relationship
-  to N sources can help managers understand the effects of alternative decisions on
-  downstream resources. N-Sink is an approach that uses widely available GIS data
-  to identify landscape sinks within HUC-12 (or larger) catchments, estimate their
-  N removal potential and map the effect of those sinks on N movement through the
-  catchment. A series of static maps are produced to visualize N removal efficiency,
-  transport and delivery, the latter in the form of an index. To facilitate data acquisition,
-  processing and visualization, an R package `nsink` was developed and applied to
-  produce static maps for all HUC-12 catchments in southern New England. The R package
-  can be used to develop N delivery maps for a specific HUC-12 of interest, or users
-  can visit the University of Connecticut website to explore previously mapped areas.
-  Users can also investigate specific flowpaths interactively by clicking on any location
-  within the catchment. A flowpath is generated, along with a table describing N removal
-  along each segment.In this software tool article we describe the motivation behind developing 
-  nsink, discuss the implementation in R, and describe the use of nsink with two use case examples. 
-  nsink is available from <https://github.com/USEPA/nsink>."
+abstract: "Background: Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water quality.
+Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery.
+Natural features within watersheds can act as landscape sinks for N, such as wetlands, streams and ponds that transform dissolved N into gaseous N, effectively removing it from the aquatic system.
+Identifying and evaluating these landscape sinks and their spatial relationship to N sources can help managers understand the effects of alternative decisions on downstream resources.
+Methods: The N-Sink approach uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment.
+Static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index.
+The R package *nsink* was developed to facilitate data acquisition, processing and visualization.
+Results: The R package creates static maps for a specific HUC-12 , or users can visit the University of Connecticut [website](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31) to explore previously mapped areas.
+Users can also investigate specific flowpaths interactively by clicking on any location within the catchment.
+A flowpath is generated, with a table describing N removal along each segment.
+We describe the motivation behind developing
+*nsink*, discuss implementation in R, and present two use case examples. *nsink* is available from <https://github.com/USEPA/nsink>.  
+Conclusions: N-Sink is a deccision support tool and created for local decision-makers to facilitate better understanding of the relationship between land use and downstream N delivery.
+Local decision-makers that have prioritized N mitigation in their long-term planning can use *nsink* to better understand the potential impact of proposed development projects and zoning variances.
+Similarly, land trusts and other NGOs interested in N mitigation can use *nsink* to identify high priority areas for acquisition or restoration."
 ---
@@
   eval = TRUE,
-  cache = TRUE,
+  cache = FALSE,
   tidy = FALSE,
@@
 library(nsink)
-library(raster)
+#library(raster)
 library(ggplot2)
@@
 
-The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] .
+The *nsink* R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] .
 Datasets used by `nsink` include the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web].
@@
 
-<!-- EPA Review Comment: If the paper is about N-Sink integrating nsink it feels like the introduction could benefit from some more mention of N-Sink’s value-add at the end of the intro.
-Even just a sentence about a pre-analyzed web map being accessible to people who aren’t familiar with R.-->
-
 Methods
@@
 
-## Development of the R package `nsink`
-The package `nsink` implements the approach detailed in Kellogg et al.
+## *Implementation in R*
+The package *nsink* implements the approach detailed in Kellogg et al.
 [-@kellogg2010geospatial] to estimate relative nitrogen (N) removal along a flowpath.
-The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder.
+The *nsink* package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder.
 The initial version required time-consuming data manipulation by hand due to limitations of earlier NHDPlus datasets as well as the limitations of working in a GIS environment requiring a user license.
-With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink was automated and applied to other HUC-12 catchments, leading to the development of the `nsink` package.
-Specifically, `nsink` relies on packages `sf` and `raster`.
-The `raster` package in R will soon be superseded by the `terra` package, which will be integrated in future `nsink` package versions.
+With the increasingly more versatile GIS packages now available in R Project for Statistical Computing (RRID:SCR_001905), the previously time-consuming spatial data acquisition and management of N-Sink was automated and applied to other HUC-12 catchments, leading to the development of the *nsink* package.
+Specifically, *nsink* relies on packages *sf* and *terra*.
 The `nsink` package is available from https://github.com/usepa/nsink and is fully described in [@hollister2022nsink].
 
-## Using the `nsink` package
-The `nsink` package provides several functions (Table 1) to set up and run an N-Sink analysis for a specified 12-digit HUC.
+## Using *nsink*
+The *nsink* package provides several functions (Table 1) to set up and run an N-Sink analysis for a specified 12-digit HUC.
 By using the package all required data are downloaded, prepared for the analysis, HUC-wide N removal calculated, and flowpaths summarized.
@@
 ```{r table1, include = TRUE, echo = FALSE}
-knitr::kable(table1, caption = "Table 1: Functions in the R package, `nsink`, with short descriptions.")
+knitr::kable(table1, caption = "Table 1: Functions in the R package, *nsink*, with short descriptions.")
 ```
@@
 ## Download the data
-The first step for an N-Sink analysis with the `nsink` package is to download the required datasets.
+The first step for an N-Sink analysis with the *nsink* package is to download the required datasets.
 The only required information to download the data is a HUC identifier.
-The `nsink` package was developed using 12-digit HUC IDs from NHDPlus, but larger HUCs (e.g.
+The *nsink* package was developed using 12-digit HUC IDs from NHDPlus, but larger HUCs (e.g.
 8-digit) may also be used.
@@
 These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC.
-The `nsink` package utilizes widely available data from several U.S.
+The *nsink* package utilizes widely available data from several U.S.
 Federal Government sources and as of `r lubridate::today()` no authentication is required to access these sources.
@@
 ```{r show-summary, include = TRUE, echo = FALSE}
-knitr::kable(niantic_fp_removal_a, caption = "Table 2: Summary of N removal along specified flowpath A (Fig. 4), as generated by the R package, nsink.")
+knitr::kable(niantic_fp_removal_a, caption = "Table 2: Summary of N removal along specified flowpath A (Fig. 4), as generated by the R package, *nsink*.")
 ```
@@
 The workflow described above includes all the basic functionality.
-Some users may wish to use `nsink` to calculate the base layers for an N-Sink analysis and then build an application outside of R.
+Some users may wish to use *nsink* to calculate the base layers for an N-Sink analysis and then build an application outside of R.
 A convenience function that downloads all data, prepares, calculates removal, and generates static maps has been included to facilitate this type of analysis.
@@
 
-Use Examples
+Operation
+============
+The R package *nsink* requires R version >= 3.5.0. The *nsink* package is available from https://github.com/usepa/nsink and is fully described in @hollister2022nsink.
+
+Use Cases
 ============
 
-## Example 1: Comparison of N delivery from two adjacent locations
+## Use Case 1: Comparison of N delivery from two adjacent locations
 
@@
 
-```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations within the Niantic River HUC-12 with differing N removal estimates.
+```{r plot4, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations within the Niantic River HUC-12 with differing N removal estimates.
 Hydrofeatures are pictured on the Nitrogen Transport Index map.", fig.height = 10, fig.width = 7.5}
@@
 
-tnsport_idx <- classify(rast(niantic_static_maps$transport_idx),matrix(c(from,to,to),ncol = 3))
+tnsport_idx <- classify(niantic_static_maps$transport_idx,matrix(c(from,to,to),ncol = 3))
 values(tnsport_idx) <- factor(values(tnsport_idx))
@@
 
-## Example 2: Source controls and sink protection or restoration
+## Use Case 2: Source controls and sink protection or restoration
 
@@
 The N Delivery Index map combines N sources with N transport to highlight developed areas that are located on "leakier" parts of a catchment.
-Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled.
+Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled.Figure 5B displays the Nitrogen Transport Index map for the same area, showing the western area to be “leakier” than the eastern area.
 
-```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (summary given in Table 4)."}
+```{r plot5a, eval = TRUE, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (summary given in Table 4)."}
 
@@
 legend_nlcd <- pal_nlcd()
-case2 <- crop(rast(niantic_data$nlcd), e)
+case2 <- crop(niantic_data$nlcd, e)
 vals <- unique(case2$Class)
-leg <- legend_nlcd[legend_nlcd$ID %in% vals$Class,]
+leg <- legend_nlcd[legend_nlcd$Class %in% vals$Class,]
 values(case2) <- factor(values(case2))
@@
 ```{r table4-build, eval = TRUE, include = FALSE}
+# extract/mask LULC values from west and east
+lu_w <- terra::mask(niantic_data$nlcd, vect(ptw_buf)) # don't use case2 as data lost when cropped
+lu_e <- terra::mask(niantic_data$nlcd, vect(pte_buf))
+
+luw_sum <- terra::freq(lu_w) %>% dplyr::select(value, count)
+names(luw_sum) <- c("value", "count_w")
+lue_sum <- terra::freq(lu_e) %>% dplyr::select(value, count)
+names(lue_sum) <- c("value", "count_e")
+
+lu_sum <- full_join(luw_sum, lue_sum)
+
+# aggregate
+dev <- c("Developed, Open Space", "Developed, Low Intensity", 
+         "Developed, Medium Intensity","Developed, High Intensity")
+ag <- c("Pasture/Hay", "Cultivated Crops")
+forest_veg <- c("Deciduous Forest", "Evergreen Forest", "Mixed Forest",
+                "Dwarf Scrub", "Shrub/Scrub", "Grassland/Herbaceous",
+                "Sedge/Herbaceous", "Lichens", "Moss")
+wetland <- c("Woody Wetlands", "Emergent Herbaceous Wetlands")
+water <- "Open Water"
+other <- c("Perennial Ice/Snow", "Barren Land (Rock/Sand/Clay)")
+
+lu_sum$agg <- "Developed"
+lu_sum$agg[lu_sum$value %in% ag] <- "Agriculture"
+lu_sum$agg[lu_sum$value %in% forest_veg] <- "Forest/Vegetated"
+lu_sum$agg[lu_sum$value %in% wetland] <- "Wetlands"
+lu_sum$agg[lu_sum$value == "Open Water"] <- "Open Water"
+lu_sum$agg[lu_sum$value %in% other] <- "Other"
+
+lu_agg <- lu_sum %>% group_by(agg) %>% 
+  summarize(sum_w = sum(count_w, na.rm = TRUE), 
+            sum_e = sum(count_e, na.rm = TRUE)) %>%
+  mutate(percent_w = round(100*sum_w/sum(sum_w, na.rm = TRUE)),
+         percent_e = round(100*sum_e/sum(sum_e, na.rm = TRUE)))
+
+dev_pos <- which(lu_agg$agg == "Developed")
+ag_pos <- which(lu_agg$agg == "Agriculture")
+for_pos <- which(lu_agg$agg == "Forest/Vegetated")
+wet_pos <- which(lu_agg$agg == "Wetlands")
+other_pos <- which(lu_agg$agg == "Other")
+
 table4 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
                                       "Wetlands", "Other"),
-                     "West (%)" = c("23","7","64","6","0"),
-                     "East (%)" = c("20","0","61","18","1"),
+                     "West (%)" = c(lu_agg$percent_w[dev_pos], lu_agg$percent_w[ag_pos],
+                                    lu_agg$percent_w[for_pos], lu_agg$percent_w[wet_pos],
+                                    lu_agg$percent_w[other_pos]),
+                     "East (%)" = c(lu_agg$percent_e[dev_pos], lu_agg$percent_e[ag_pos],
+                                    lu_agg$percent_e[for_pos], lu_agg$percent_e[wet_pos],
+                                    lu_agg$percent_e[other_pos]),
                      check.names = FALSE)
 ```
+
 ```{r table4, include = TRUE, echo = FALSE}
@@
 
-```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Transport Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas."}
+```{r, plot5b, eval = TRUE, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Transport Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas."}
 
@@
 
-tnsport_idx <- classify(rast(niantic_static_maps$transport_idx),matrix(c(from,to,to),ncol = 3))
+tnsport_idx <- classify(niantic_static_maps$transport_idx,matrix(c(from,to,to),ncol = 3))
 case2_tnsport <- crop(tnsport_idx, e)
@@
 to <- c(seq(10,100,10), NA)
-delivery_idx <- classify(rast(niantic_static_maps$delivery_idx),matrix(c(from,to,to),ncol = 3))
+delivery_idx <- classify(niantic_static_maps$delivery_idx,matrix(c(from,to,to),ncol = 3))
 case2_delivery <- crop(delivery_idx, e)
@@
 
-Conclusions
+Discussion
 ==========================
 
-N-Sink as implemented in the `nsink` R package allows users to easily download all necessary spatial data (hydrography, land cover and soils), create static maps (N Removal Efficiency, N Transport Efficiency and N Delivery Index), and interactively investigate N removal along flowpaths within any specified catchment.
+N-Sink as implemented in the *nsink* R package allows users to easily download all necessary spatial data (hydrography, land cover and soils), create static maps (N Removal Efficiency, N Transport Efficiency and N Delivery Index), and interactively investigate N removal along flowpaths within any specified catchment.
 Currently, the default catchment definition uses HUC-12 designations, though it is also possible to use larger catchments.
@@
 Similarly, land trusts and other NGOs interested in N mitigation can use this tool to identify high priority areas for acquisition or restoration.
+
+Data availability
+=================
+Datasets used by nsink include the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Landcover Dataset (NLCD) Impervious Surface [@jin2019overall; @moore2019user; @soil2017web].
+All are available via the FedData package in R [@feddata].
+
+Software availability
+=================
+Software available from: https://github.com/usepa/nsink
+
+Source code available from: https://github.com/usepa/nsink
+
+Archived source code at the time of publication: DOI: 10.5281/zenodo.10045141
+
+License: MIT
+
+Competing interests
+==================
+The authors declare that they have no competing interests.
+
+Grant information
+=================
+The research was supported, in part, by a cooperative agreement with the U.S.
+Environmental Protection Agency's Office of Research and Development, EPA Assistant Agreement Number CR83680001, "N-Sink tool and outreach to two watersheds in New England."
 
@@
 the U.S. Government or the U.S. Environmental Protection Agency. The EPA does 
-not endorse any commercial products, services, or enterprises. This research has 
-been supported, in part, by a cooperative agreement with the U.S. Environmental
-Protection Agency's Office of Research and Development, EPA Assistant Agreement
-Number CR83680001, "N-Sink tool and outreach to two watersheds in New England."
-This contribution is identified by the tracking number ORD-054342 of the 
-Atlantic Coastal Environmental Sciences Division, Office of Research and Development, 
-Center for Environmental Measurement and Modeling, US Environmental Protection Agency.
+not endorse any commercial products, services, or enterprises.
+This contribution is identified by the tracking number ORD-054342 of the Atlantic Coastal Environmental Sciences Division, Office of Research and Development, Center for Environmental Measurement and Modeling, US Environmental Protection Agency.
 

### 2024-05-03 manuscript/manuscript.Rmd
fixed Zenodo DOI

--- 2023-10-31 manuscript/manuscript.Rmd
@@
 Here we describe the development of the package, and example workflow, and simple use applications to demonstrate the capabilities of the tools.
-The code for this analysis is available from <https://github.com/USEPA/nsink_manuscript/> and is archived at <https://doi.org/10.5281/zenodo.8325360>.
+The code for this analysis is available from <https://github.com/USEPA/nsink_manuscript/> and is archived at <https://doi.org/10.5281/zenodo.10045141>.
 All analyses were conducted with R version 4.2.2 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2022; @here; @ggplot2; @sf1; @sf2; @knitr1; @knitr2; @knitr3; @feddata; @hollister2022nsink; @units; @raster; @terra; @tidyterra; @lubridate; @units].
@@
 
-```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 1) for the Niantic River HUC-12, showing the estimated fraction of N removed within three types of landscape N sinks - wetlands, streams, lakes/ponds/reservoirs.", echo=FALSE}
+```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 1) for the Niantic River HUC-12, showing the estimated percentage of N removed within three types of landscape N sinks - wetlands, streams, lakes/ponds/reservoirs.", echo=FALSE}
 
@@
 N loading estimates are based on land cover classes, making use of peer-reviewed published data, and normalized, with a range of 0 to 1.
-These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the N Delivery Index.", echo=FALSE}
+These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the NDI.", echo=FALSE}
 
@@
 
-```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (summary given in Table 4)."}
+```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (Table 4)."}
 
@@
 ```{r table4-build, eval = TRUE, include = FALSE}
-table4 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
+table3 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
                                       "Wetlands", "Other"),
@@
 ```{r table4, include = TRUE, echo = FALSE}
-knitr::kable(table4, caption = "Table 4: Summary of land cover in highlighted areas of interest.")
+knitr::kable(table3, caption = "Table 4: Summary of land cover in highlighted areas of interest.")
 ```

### 2024-05-03 manuscript/manuscript.Rmd
oops...  now here is the conflict fix!

--- 2024-03-06 manuscript/manuscript.Rmd
@@
 
-The *nsink* R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] .
+The `nsink` R package was written to simplify the acquisition and management of the spatial data necessary to estimate N removal within each identified landscape sink, estimate cumulative N removal along a specified flowpath, and create a set of static watershed maps [@hollister2022nsink] .
 Datasets used by `nsink` include the National Hydrography Dataset Plus (NHDPlus), Soil Survey Geographic Database (SSURGO), the National Land Cover Dataset (NLCD) land cover and the National Land Cover Dataset (NLCD) impervious surface [@jin2019overall; @moore2019user; @soil2017web].
